---
title: "Target Markdown"
output:
  html_document: default
  pdf_document: default
---
## Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(targets)
tar_unscript()
```

```{targets global-imports, tar_globals=TRUE}
library(targets)
library(tidyverse)
library(glue)
library(tarchetypes)
library(conflicted)
library(gseasusie)


#library(clustermq)
#options(clustermq.scheduler = "multiprocess")

library(future)
library(future.callr)
plan(callr)

# Manage naming conflicts
conflict_prefer("select", "dplyr", "AnnotationDbi")
conflict_prefer("filter", "dplyr")
conflict_prefer("rename", "dplyr")
conflict_prefer("lag", "dplyr")
conflict_prefer("desc", "dplyr")

# callr cant find these when used in 'tar_option_set'
# but these imports would be better placed there
# Both installed via bioconductor so maybe that's a hint
require(GSEABenchmarkeR, quietly = T)
require(EnrichmentBrowser, quietly = T)

source('code/target_factories.R')
source('code/target_simulate_gene_lists.R')
source('code/target_score_simulation.R')
source('code/load_data.R')

# options(clustermq.scheduler = "slurm", clustermq.template="clustermq_slurm.sh")
tar_option_set(workspace_on_error = TRUE)

tar_option_set(packages = c(
  "tidyverse", "WebGestaltR",
  "Matrix", "susieR", "glmnet"))
tar_option_set(memory='transient', garbage_collection = T)

REP <- tar_toggle(1, 10)
BATCH <- tar_toggle(1, 1)
```

## Methods

To add a method add a row to `method_lookup` and `score_lookup`
`method_function` should take arguments `X` and `y`
`score_function` should take arguments `fit` and `sim`

```{targets method_bank, tar_globals=TRUE}
fit_logistic_susie_l1 = function(X, y){
  gseasusie::fit_logistic_susie(X, y, L=1, maxit=500)
}
fit_linear_susie_l1 = function(X, y){
  susieR::susie(X, y, L=1)
}
fit_logistic_susie_l10 = function(X, y){
  gseasusie::fit_logistic_susie(X, y, L=10, maxit=500)
}
fit_logistic_susie_veb_boost_l10 = function(X, y){
  gseasusie::fit_logistic_susie_veb_boost(X, y, L=10, maxit=500)
}
fit_linear_susie_l10 = function(X, y){
  susieR::susie(X, y, L=10)
}
fit_ora2 = function(X, y){
  gseareg::fit_ora(X, y)
}
fit_linear_lasso = function(X, y){
  glmnet::glmnet(X, y, family='gaussian', alpha=1)
}
fit_logistic_lasso = function(X, y){
  # hack to make it not fail for rare simulations
  if(sum(y) < 2){
    y[sample(length(y), 2)] <- 1
  } else if (sum(y) > length(y) - 2){
    y[sample(length(y), 2)] <- 0
  }
  glmnet::glmnet(X, y, family='binomial', alpha=1)
}

method_lookup <- tribble(
  ~method_name, ~method_function, ~method_args,
  'ora', gseasusie::fit_ora, list(),
  'logistic_susie.l1', fit_logistic_susie_l1, list(),
  'linear_susie.l1', fit_linear_susie_l1, list(),
  'logistic_susie.l10', fit_logistic_susie_l10, list(),
  'logistic_susie_veb_boost.l10', fit_logistic_susie_veb_boost_l10, list(),
  'linear_susie.l10', fit_linear_susie_l10, list(),
  'linear_lasso', fit_linear_lasso, list(),
  'logistic_lasso', fit_logistic_lasso, list()
)

score_lookup <- tribble(
  ~method_name, ~score_function,
  'ora', score_phyper,
  'logistic_susie.l1', score_pip,
  'linear_susie.l1', score_pip,
  'logistic_susie.l10', score_pip,
  'logistic_susie_veb_boost.l10', score_pip,
  'linear_susie.l10', score_pip,
  'linear_lasso', score_glm,
  'logistic_lasso', score_glm
)
```

## Targets

### Get gene sets
```{targets load-genesets}
list(
  tar_target(genesets, gseasusie::load_gene_sets()),
  tar_target(X.gonr, genesets$gobp_nr$X)
)
```

## L1 Simulation

Simulate a single active gene set from `GO-BP no redundnant`

```{targets l1.sim}
## set simulation parameters
get.l1.sim.params <- function(){
  params <- crossing(
    background.logit=c(-8, -6, -4, -2),
    active.logit=c(-4, -2, 0, 2, 4),
    sim_function = c('simulate.constant.sim'),
    X='X.gonr')

  simulate.constant.sim
  params <- params %>% mutate(
    param_names = list((params %>% select(!c(X, sim_function)) %>% colnames()))
  ) %>% rowwise() %>% mutate(
    params = list(setNames(c_across(!c(X, sim_function, param_names)), param_names))
  )
}

# target metadata/specification
l1.gonr.sim.params <- get.l1.sim.params()
l1.gonr.methods <- method_lookup
l1.gonr.score_lookup <- l1.gonr.methods %>%
  left_join(score_lookup)

# execute pipeline
list(
  make_sim_spec('l1.gonr', l1.gonr.sim.params, BATCH, REP),
  sim_factory('l1.gonr', X.gonr, l1.gonr.spec),
  fit_factory('l1.gonr', X.gonr, l1.gonr.sim, l1.gonr.methods),
  score_factory('l1.gonr', l1.gonr.score_lookup),
  aggegate_score_factory('l1.gonr', l1.gonr.score_lookup)
)
```

```{targets additive.l5.sim}
## set simulation parameters
get.additive.l5.sim.params <- function(){
  params <- crossing(
    background.logit=seq(-8, -2, 2),
    active.logit=seq(2, 6, 1),
    sim_function = 'simulate.constant.additive.sim',
    L = c(5),
    X='X.gonr')
  params <- params %>% mutate(
    param_names = list((params %>% select(!c(X, sim_function)) %>% colnames()))
  ) %>% rowwise() %>% mutate(
    params = list(setNames(c_across(!c(X, sim_function, param_names)), param_names))
  )
  return(params)
}

# target metadata/specification
additive.l5.gonr.sim.params <- get.additive.l5.sim.params()
additive.l5.gonr.methods <- method_lookup
additive.l5.gonr.score_lookup <- additive.l5.gonr.methods %>%
  left_join(score_lookup)

# execute pipeline
list(
  make_sim_spec('additive.l5.gonr', additive.l5.gonr.sim.params, BATCH, REP),
  sim_factory('additive.l5.gonr', X.gonr, additive.l5.gonr.spec),
  fit_factory('additive.l5.gonr', X.gonr, additive.l5.gonr.sim, additive.l5.gonr.methods),
  score_factory('additive.l5.gonr', additive.l5.gonr.score_lookup),
  aggegate_score_factory('additive.l5.gonr', additive.l5.gonr.score_lookup)
)
```

```{r tar_make, eval=FALSE}
tar_make()
```

```{r}
knitr::knit_exit()
```

## L1 Sim plots

```{r}
l1.sim.aggregate.scores <- read_tsv('l1.sim.aggregate.scores')

# TPR vs FPR
plotdat <- l1.sim.aggregate.scores %>%
  filter(!(method_name %in% c('linear_lasso', 'logistic_lasso'))) %>%
  filter(active.logit > -4, thresh > 0.05, thresh < 0.999) %>%
  group_by(method_name, thresh) %>%
  summarise(TP = sum(TP), FP = sum(FP), TN = sum(TN), FN = sum(FN)) %>%
  mutate(TPR = TP/(TP + FN), FDP  = FP/(TP + FP)) 

plotdat %>%
  filter(method_name == 'logistic_susie.l10') %>%
  arrange(desc(thresh))

plotdat %>%
  ggplot(aes(x=FDP, y=TPR, color=method_name)) + 
  geom_line()
```


```{r}
a <- l1.sim.aggregate.scores %>%
  filter(method_name != 'ora', near(thresh, 0.95)) %>%
  group_by(method_name, background.logit, active.logit) %>%
  summarise(TP = sum(TP), FP = sum(FP), TN = sum(TN), FN = sum(FN)) %>%
  mutate(
    recall = TP / (TP + FN),
    precision = TP / (TP + FP),
    specificity = TN / (TN + FP)
  ) %>%
  pivot_longer(c(recall, precision, specificity), names_to= 'metric') 

b <- l1.sim.aggregate.scores %>%
  filter(method_name == 'ora', near(thresh, 0.999, 1e-10)) %>%
  group_by(method_name, background.logit, active.logit) %>%
  summarise(TP = sum(TP), FP = sum(FP), TN = sum(TN), FN = sum(FN)) %>%
  mutate(
    recall = TP / (TP + FN),
    precision = TP / (TP + FP),
    specificity = TN / (TN + FP)
  ) %>%
  pivot_longer(c(recall, precision, specificity), names_to= 'metric') 

bind_rows(a, b) %>%
  ggplot(aes(x=background.logit, y=active.logit, fill=value)) + 
  geom_tile() + facet_grid(vars(metric), vars(method_name)) + theme(aspect.ratio = 1)
  ggplot(aes(x=background.logit, y=active.logit, fill=TPR)) + geom_tile()
```


## L5 Sim plots

```{r}
scores <- read_rds('additive.l5.gonr.aggregate.scores')
scores <- scores %>% unnest(scores)

# TPR vs FPR
plotdat <- scores %>%
  filter(!(method_name %in% c('linear_lasso', 'logistic_lasso'))) %>%
  filter(thresh > 0.05, thresh < 0.999) %>%
  group_by(method_name, thresh) %>%
  summarise(TP = sum(TP), FP = sum(FP), TN = sum(TN), FN = sum(FN)) %>%
  mutate(TPR = TP/(TP + FN), FDP  = FP/(TP + FP)) 

plotdat %>%
  filter(method_name == 'logistic_susie.l10') %>%
  arrange(desc(thresh))

plotdat %>%
  ggplot(aes(x=FDP, y=TPR, color=method_name)) + 
  geom_line()
```


```{r}
plotdat <- scores %>%
  filter(!(method_name %in% c('linear_lasso', 'logistic_lasso'))) %>%
  filter(near(thresh, 0.99)) %>%
  group_by(method_name, background.logit, active.logit) %>%
  summarise(TP = sum(TP), FP = sum(FP), TN = sum(TN), FN = sum(FN)) %>%
  mutate(
    recall = TP / (TP + FN),
    precision = TP / (TP + FP),
    specificity = TN / (TN + FP)
  ) %>%
  pivot_longer(c(recall, precision, specificity), names_to= 'metric') 


plotdat %>%
  ggplot(aes(x=background.logit, y=active.logit, fill=value)) + 
  geom_tile() + facet_grid(vars(metric), vars(method_name)) + theme(aspect.ratio = 1)
  ggplot(aes(x=background.logit, y=active.logit, fill=TPR)) + geom_tile()
```


```{targets l1.sim.old}
## pick methods to include in this sim
l1.gonr.methods <- method_bank %>% dplyr::filter(
  method.name %in% c(
    'fet',
    'susie.10',
    'logistic.susie.10',
    'logistic.susie.veb.boost.10',
    'lasso.gaussian',
    'elastic.net.gaussian',
    'lasso.binomial',
    'elastic.net.binomial',
    'mr.ash.lasso'
  )) %>%
  mutate(ma_names = map(method_args, names))

list(
  make_sim_spec('l1.gonr', l1.gonr.sim.params, BATCH, REP),
  sim_factory('l1.gonr', X.gonr, l1.gonr.spec),
  fit_factory('l1.gonr', X.gonr, l1.gonr.sim, l1.gonr.methods),
  score_factory('l1.gonr', l1.gonr.methods)
)
```



```{targets constant.sim}
## set simulation parameters
get.constant.sim.params <- function(){
  params <- crossing(
    background.logit=c(-8, -6, -4, -2),
    active.logit=c(-2, -1, 0, 1, 2),
    sim_function = c('simulate.constant.sim'),
    L = c(5, 10, 20),
    X='X.gonr')
  params <- params %>% mutate(
    param_names = list((params %>% select(!c(X, sim_function)) %>% colnames()))
  ) %>% rowwise() %>% mutate(
    params = list(setNames(c_across(!c(X, sim_function, param_names)), param_names))
  )
}
constant.gonr.sim.params <- get.constant.sim.params()

## pick methods to include in this sim
constant.gonr.methods <- method_bank %>% dplyr::filter(
  method.name %in% c(
    'fet',
    'susie.10',
    'susie.20',
    'logistic.susie.10',
    'logistic.susie.veb.boost.10',
    'logistic.susie.veb.boost.20',
    'lasso.gaussian',
    'elastic.net.gaussian',
    'lasso.binomial',
    'elastic.net.binomial'
  )) %>%
  mutate(ma_names = map(method_args, names))

list(
  make_sim_spec('constant.gonr', constant.gonr.sim.params, BATCH, REP),
  sim_factory('constant.gonr', X.gonr, constant.gonr.spec),
  fit_factory('constant.gonr', X.gonr, constant.gonr.sim, constant.gonr.methods),
  #score_factory('constant.gonr', constant.gonr.methods)
)
```




## Constant Additive Sim

Fixed background rate, fixed beta across $L$ active gene sets.
The point of this simulation is to explore the relative performance of
SuSiE and logistic SuSiE across a range of enrichment/background levels

```{targets constant.additive.sim}
## set simulation parameters
get.constant.additive.sim.params <- function(){
  params <- crossing(
    background.logit=seq(-10, -2, 2),
    active.logit=c(0:10),
    sim_function = 'simulate.constant.additive.sim',
    L = c(1, 5, 10, 20),
    X='X.gonr') %>%
    filter(((background.logit + active.logit) >= 0)) %>%
    filter((background.logit + active.logit <= 5))

  params <- params %>% mutate(
    param_names = list((params %>% select(!c(X, sim_function)) %>% colnames()))
  ) %>% rowwise() %>% mutate(
    params = list(setNames(c_across(!c(X, sim_function, param_names)), param_names))
  )
}
constant.additive.gonr.sim.params <- get.constant.additive.sim.params()

## pick methods to include in this sim
constant.additive.gonr.methods <- method_bank %>% dplyr::filter(
  method.name %in% c(
    'fet',
    'susie.10',
    'susie.20',
    'logistic.susie.10',
    'logistic.susie.veb.boost.10',
    'logistic.susie.veb.boost.20',
    'lasso.gaussian',
    'elastic.net.gaussian',
    'lasso.binomial',
    'elastic.net.binomial',
    'mr.ash.alpha'
  )) %>%
  mutate(ma_names = map(method_args, names))

list(
  make_sim_spec('constant.additive.gonr', constant.additive.gonr.sim.params, BATCH, REP),
  sim_factory('constant.additive.gonr', X.gonr, constant.additive.gonr.spec),
  fit_factory('constant.additive.gonr', X.gonr, constant.additive.gonr.sim, constant.additive.gonr.methods),
  score_factory('constant.additive.gonr', constant.additive.gonr.methods)
)
```


## Summarise

```{r}

```



## Real Data

```{targets pbmc.deseq}
run_enrichment = function(db, experiment, thresh, genesets, data){
  message(paste0('Fitting logistic susie...',
    '\n\tExperiment = ', experiment,
    '\n\tDatabase = ', db,
    '\n\tthresh = ', thresh))
  
  bin.data <- gseasusie::prep_binary_data(genesets[[db]], data[[experiment]], thresh)
  X <- bin.data$X
  y <- bin.data$y
  
  logistic.fit <- gseasusie::fit_logistic_susie_veb_boost(X, y, L=20)
  ora <- gseasusie::fit_ora(X, y)
  
  linear.fit <- susieR::susie(X, y)
  marginal_regression <- gseasusie::fit_marginal_regression_jax(X, y)
  residual_regression <- gseasusie::fit_residual_regression_jax(X, y, logistic.fit)
  
  res <- tibble(
    db = db,
    experiment = experiment,
    thresh = thresh,
    fit=list(logistic.fit),
    linear.fit = list(linear.fit),
    ora=list(ora),
    marginal_reg = list(marginal_regression),
    residual_reg = list(residual_regression)
  )
  return(res)
}

summarise_enrichment = function(fit, ora, marginal_reg, residual_reg, geneSetDes){
  # put everything into one table
  gssum <- get_gene_set_summary(fit)
  cssum <- get_cs_summary_condensed(fit)
  
  reg <- left_join(
    marginal_reg,
    residual_reg,
    by="geneSet", suffix=c('_marginal', '_residual'))

  gene_set_summary <- 
    gssum %>%
    dplyr::left_join(cssum) %>%
    dplyr::left_join(ora) %>%
    left_join(reg) %>%
    left_join(geneSetDes)
  return(gene_set_summary)
}

GENESETS <- c('c1')
THRESHOLDS <- c(1e-4) # c(1e-2, 1e-4, 1e-6, 1e-8)
PREFIX <- 'sc_pbmc_deseq'
DATALOADER <- load_sc_pbmc_deseq2

list(
  # get gene sets
  tar_target(genesets, gseasusie::load_gene_sets(GENESETS)),
  # get data
  tar_target(sc_pbmc_deseq_data, DATALOADER()),
  # make meta table
  tar_target(sc_pbmc_deseq_meta, crossing(
    db=names(genesets),
    experiment=names(sc_pbmc_deseq_data),
    threshold = THRESHOLDS)),
  # fit enrichment for each row of meta
  tar_target(
    sc_pbmc_deseq_enrichment_fit,
    run_enrichment(
      sc_pbmc_deseq_meta$db,
      sc_pbmc_deseq_meta$experiment,
      sc_pbmc_deseq_meta$threshold,
      genesets,
      sc_pbmc_deseq_data),
    pattern = map(sc_pbmc_deseq_meta)
  ),
  # summarise enrichment (lightweight, used for plotting, etc)
  tar_target(
    sc_pbmc_deseq_enrichment_summary,
    sc_pbmc_deseq_enrichment_fit %>%
      rowwise() %>% 
      mutate(
        enrichment_summary = list(summarise_enrichment(
          fit, ora, marginal_reg, residual_reg, genesets[[db]]$geneSetDes))) %>%
      select(db, experiment, thresh, enrichment_summary),
    pattern = map(sc_pbmc_deseq_enrichment_fit)
  )
)
```

## Pipeline

If you ran all the `{targets}` chunks in non-interactive mode, then your R scripts are set up to run the pipeline.

```{r}
#tar_make()
```

The `targets` dependency graph helps your readers understand the steps of your pipeline at a high level.

```{r}
tar_visnetwork()
```

At this point, you can go back and run `{targets}` chunks in interactive mode without interfering with the code or data of the non-interactive pipeline.
