---
title: "Target Markdown"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(targets)
tar_unscript()
options(clustermq.scheduler = "multiprocess")
```

```{targets globals, tar_globals=TRUE}
library(tidyverse)
library(tidyr)
library(glue)
library(tarchetypes)
library(clustermq)
library(conflicted)

# Manage naming conflicts

conflict_prefer("select", "dplyr", "AnnotationDbi")
conflict_prefer("filter", "dplyr")
conflict_prefer("rename", "dplyr")
conflict_prefer("lag", "dplyr")
conflict_prefer("desc", "dplyr")

# callr cant find these when used in 'tar_option_set'
# but these imports would be better placed there
# Both installed via bioconductor so maybe that's a hint
require(GSEABenchmarkeR, quietly = T)
require(EnrichmentBrowser, quietly = T)

source('code/load_gene_sets.R')
source('code/simulate_gene_lists.R')
source('code/score_simulations.R')
source('target_components/methods.R')
source('target_components/factories.R')

options(clustermq.scheduler = "multiprocess")
tar_option_set(workspace_on_error = TRUE)

tar_option_set(packages = c(
  "tidyverse", "WebGestaltR",
  "Matrix", "susieR", "glmnet",
  "mr.ash.alpha", "VEB.Boost", "fastglm"))
tar_option_set(memory='transient', garbage_collection = T, error='workspace')


get.test.sim.params <- function(){
  return(crossing(
    active.logit=c(-6, -5, -4, -3, -2),
    background.logit=c(-2, -1, 0, 1, 2),
    sim_function = c('simulate.constant.sim')))
}
test.sim.params <- get.test.sim.params()
test.sim.params <- test.sim.params %>% rowwise() %>%
  mutate(
    params = list(
      setNames(
        c_across(!sim_function),
        colnames(select(test.sim.params, !sim_function)))),
    X='X.gonr'
  )

get.methods <- function(){
  method_bank %>% dplyr::filter(
    method.name %in% c(
      'fet',
      'susie.10',
      'logistic.susie.veb.boost.10'
    )
  )
}
methods <- get.methods()


REP <- 1
BATCH <- tar_toggle(1, 10)
```

Near the top of the document, you may also wish to remove the `_targets_r` directory previously written by non-interactive runs of the report. Otherwise, your pipeline may contain superfluous targets.


## Get gene sets
```{targets load-genesets}
tar_target(X.gonr, load.webGestalt.X())
```


## Simulations

## L1 Sim

Simulate a single active gene set from `GO-BP no redundnant`

```{targets l1.sim}
## set simulation parameters
get.l1.sim.params <- function(){
  params <- crossing(
    background.logit=c(-8, -6, -4, -2),
    active.logit=c(-2, -1, 0, 1, 2),
    sim_function = c('simulate.constant.sim'),
    X='X.gonr')

  params <- params %>% mutate(
    param_names = list((params %>% select(!c(X, sim_function)) %>% colnames()))
  ) %>% rowwise() %>% mutate(
    params = list(setNames(c_across(!c(X, sim_function, param_names)), param_names))
  )
}
l1.gonr.sim.params <- get.l1.sim.params()

## pick methods to include in this sim
l1.gonr.methods <- method_bank %>% dplyr::filter(
  method.name %in% c(
    'fet',
    'susie.10',
    'logistic.susie.veb.boost.10'
  )) %>%
  mutate(ma_names = map(method_args, names))


list(
  sim_factory('l1.gonr', X.gonr, l1.gonr.sim.params, BATCH, REP),
  fit_factory('l1.gonr', X.gonr, l1.gonr.sim, l1.gonr.methods),
  score_factory('l1.gonr', l1.gonr.methods)
)
```


```{targets constant.sim}
## set simulation parameters
get.constant.sim.params <- function(){
  params <- crossing(
    background.logit=c(-8, -6, -4, -2),
    active.logit=c(-2, -1, 0, 1, 2),
    sim_function = c('simulate.constant.sim'),
    L = c(5, 10, 20),
    X='X.gonr')
  params <- params %>% mutate(
    param_names = list((params %>% select(!c(X, sim_function)) %>% colnames()))
  ) %>% rowwise() %>% mutate(
    params = list(setNames(c_across(!c(X, sim_function, param_names)), param_names))
  )
}
constant.gonr.sim.params <- get.constant.sim.params()

## pick methods to include in this sim
constant.gonr.methods <- method_bank %>% dplyr::filter(
  method.name %in% c(
    'fet',
    'susie.10', 'susie.30',
    'logistic.susie.veb.boost.10', 'logistic.susie.veb.boost.30'
  )) %>%
  mutate(ma_names = map(method_args, names))


list(
  sim_factory('constant.gonr', X.gonr, constant.gonr.sim.params,  BATCH, REP),
  fit_factory('constant.gonr', X.gonr, constant.gonr.sim, constant.gonr.methods),
  score_factory('constant.gonr', constant.gonr.methods)
)
```

## Constant Additive Sim

Fixed background rate, fixed beta across $L$ active gene sets.
The point of this simulation is to explore the relative performance of
SuSiE and logistic SuSiE across a range of enrichment/background levels

```{targets constant.additive.sim}
## set simulation parameters
get.constant.additive.sim.params <- function(){
  params <- crossing(
    background.logit=seq(-10, -2, 2),
    active.logit=(0:10),
    sim_function = c('simulate.constant.additive.sim'),
    L = c(1, 5, 10, 20),
    X='X.gonr') %>%
    filter(((background.logit + active.logit) >= 0)) %>%
    filter((background.logit + active.logit <= 5))

  params <- params %>% mutate(
    param_names = list((params %>% select(!c(X, sim_function)) %>% colnames()))
  ) %>% rowwise() %>% mutate(
    params = list(setNames(c_across(!c(X, sim_function, param_names)), param_names))
  )
}
constant.additive.gonr.sim.params <- get.constant.additive.sim.params()

## pick methods to include in this sim
constant.additive.gonr.methods <- method_bank %>% dplyr::filter(
  method.name %in% c(
    'fet',
    'susie.10', 'susie.30',
    'logistic.susie.veb.boost.10', 'logistic.susie.veb.boost.30'
  )) %>%
  mutate(ma_names = map(method_args, names))

list(
  sim_factory('constant.additive.gonr', X.gonr, constant.additive.gonr.sim.params,  BATCH, REP),
  fit_factory('constant.additive.gonr', X.gonr, constant.additive.gonr.sim, constant.additive.gonr.methods),
  score_factory('constant.additive.gonr', constant.additive.gonr.methods)
)
```
## Pipeline

If you ran all the `{targets}` chunks in non-interactive mode, then your R scripts are set up to run the pipeline.

```{r}
#tar_make()
```

The `targets` dependency graph helps your readers understand the steps of your pipeline at a high level.

```{r}
tar_visnetwork()
```

At this point, you can go back and run `{targets}` chunks in interactive mode without interfering with the code or data of the non-interactive pipeline.
