---
title: "alpha_for_single_cell"
author: "karltayeb"
date: "2022-04-09"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

## Introduction

Most GSEA methods operate on z-score/p-values
Methods like `ash` use effect sizes and standard errors
Both make implicit assumptions about the relationship between standard error and effect size. z-score model implies that larger standard errors imply larger effects on average. The summary stat model assumes standard errors and effect sizes are independent.

A family of


$$
\hat \beta_i | s_i \sim \mathcal{N}(\beta, s_i^2)\\
\beta_i | s_i \sim \pi_0 \delta_0 + \sum_k \pi_k \mathcal{N}(0, s_i^{2\alpha} \sigma^2_{0k})
$$


Before getting into the enrichment problem, we can explore how well the data support different values of
$\alpha$. Peter's already shown for the droplet DE data that when we fit ASH with a grid of $\alpha$ values we strongly prefer $\alpha=0$. That is reproduced below.

We also carry out the same analysis with a point normal (where the normal component has variation $\sigma_i = s_i^\alpha \sigma_0$). For the point normal we estimate $\pi_0$ and $\sigma_0$ across a grid of $\alpha$ values.

While the point normal also prefers small $\alpha$ I think it is not really behaving the way we want to. Almost all the shrinkage from $\alpha=1$ model comes from the point mass, which almost exclusively emphasizes tests with small standard errors. Almost all the shrinkage from $\alpha=0$ comes from small estimate of $\hat \sigma_0$, which could cause overly aggressive shrinkage of large effects.


```{r load.data}
load("~/Research/logistic-susie-gsea/data/de-droplet/de-droplet-noshrink.RData")
```


## Fit ASH and Point normal

```{r fit.ash}
library(fastTopics)
library(ashr)
library(ggplot2)
library(cowplot)
library(tidyverse)
set.seed(1)

# Load the results of the DE analysis.
load("data/de-droplet/de-droplet-noshrink.RData")
b  <- c(de_merged$postmean)
se <- c(with(de_merged,postmean/z))

#b <- de_merged$postmean[, 1]
#se <- de_merged$postmean[,1] / de_merged$z[,1]

i  <- which(!is.na(se))
b  <- b[i]
se <- se[i]

# Run adaptive shrinkage for different settings of alpha.
a <- seq(0,1,0.05)
n <- length(a)
ash.fits <- xfun::cache_rds({
  vector("list",n)
  for (i in 1:n) {
    cat(i,"")
    fits[[i]] <- ash(b,se,alpha = a[i])
  }
  cat("\n")
  fits
}, dir='cache/alpha_for_single_cell', file='ash.fits')
```

```{r fit.point.normal}
source('code/marginal_sumstat_gsea.R')

fit.fixed.alpha = function(a){
  cat(a,"")
  x <- rbinom(length(b), 1, 0.5)
  gsea <- summary.stat.gsea(x=x, beta=b.mod, se=se)
  params.init = list(
    theta = c(0, 0),
    alpha = a,
    sigma0 = 1.0
  )
  res <- gsea$expectation.maximiztion(params.init, tol=1e-6, update.theta1 = F, update.alpha = F)
  res$loglik <- tail(res$lik.history, 1)
  
  # compute posterior mean
  pi1.post <- res$responsibilities
  var.post <- 1 / (1/se^2 + 1/res$sigma0^2)
  mu.post  <- var.post * b / se^2
  res$PosteriorMean <- mu.post * pi1.post
  return(res)
}
point.normal.fits <- xfun::cache_rds({
  map(a, ~ fit.fixed.alpha(.x))
}, dir='cache/alpha_for_single_cell', file='point.normal.fits')
```


```{r collect.results}
pdat <- data.frame(
  beta.hat = b,
  se = se,
  beta.post.ash.0 = ash.fits[[1]]$result$PosteriorMean,
  beta.post.ash.1 = ash.fits[[n]]$result$PosteriorMean,
  beta.post.pn.0 = point.normal.fits[[1]]$PosteriorMean,
  resp.pn.0 = point.normal.fits[[1]]$responsibilities,
  beta.post.pn.1 = point.normal.fits[[n]]$PosteriorMean,
  resp.pn.1 = point.normal.fits[[n]]$responsibilities
)

q.ash.0 = quantile(ash.fits[[1]]$result$PosteriorMean, seq(0.05,0.95,0.01))
q.ash.1 = quantile(ash.fits[[n]]$result$PosteriorMean, seq(0.05,0.95,0.01))
q.pn.0 = quantile(point.normal.fits[[1]]$PosteriorMean, seq(0.05,0.95,0.01))
q.pn.1 = quantile(point.normal.fits[[n]]$PosteriorMean, seq(0.05,0.95,0.01))
quantiles <- as.data.frame(cbind(q.ash.0, q.ash.1, q.pn.0, q.pn.1))

loglik.ash <- map_dbl(ash.fits, ~pluck(.x, 'loglik'))
loglik.pn <- map_dbl(point.normal.fits, ~pluck(.x, 'loglik'))
loglik <- data.frame(
  alpha=a,
  loglik.ash = loglik.ash,
  loglik.point.normal = loglik.pn
)

pi0.ash <- map_dbl(ash.fits, ~pluck(.x, 'fitted_g', 'pi', 1))
theta0.pn <- map_dbl(point.normal.fits, ~pluck(.x, 'theta', 1))
sigma0.pn <- map_dbl(point.normal.fits, ~pluck(.x, 'sigma0'))
params <- data.frame(
  alpha=a,
  pi0.ash = pi0.ash,
  pi0.pn = 1 - 1/(1+exp(-theta0.pn)),
  sigma0.pn = sigma0.pn
)
```


```{r ash.plot}
# Plot likelihood vs. alpha.
p1 <- ggplot(loglik, aes(x = alpha,y = loglik.ash)) +
  geom_point() +
  geom_line() +
  ggtitle("loglik vs. alpha") +
  theme_cowplot(font_size = 12)

# Plot posterior estimates for best fit vs. worst fit.
p2 <- ggplot(pdat, aes(x = beta.post.ash.0, y = beta.post.ash.1, color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(
    low = "lightskyblue",mid = "gold", high = "orangered", midpoint = 3) +
  labs(
    x = "alpha = 0",
    y = "alpha = 1",
    color = "s.e.",
    title = "posterior estimates") +
  theme_cowplot(font_size = 12)

# Compare the distributions of the estimated coefs.

x <- with(quantiles,c(q.ash.0,q.ash.1))
p3 <- ggplot(quantiles, aes(x = q.ash.0,y = q.ash.1)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "skyblue", linetype = "dashed") +
  coord_cartesian(xlim = range(x), ylim = range(x)) +
  labs(x = "alpha = 0",y = "alpha = 1",title = "quantiles") +
  theme_cowplot(font_size = 12)
plot_grid(p1,p2,p3,nrow = 1,ncol = 3)
```


```{r point.normal.plot}
# Plot likelihood vs. alpha.
p1 <- ggplot(loglik, aes(x = alpha,y = loglik.point.normal)) +
  geom_point() +
  geom_line() +
  ggtitle("loglik vs. alpha") +
  theme_cowplot(font_size = 12)

# Plot posterior estimates for best fit vs. worst fit.
p2 <- ggplot(pdat, aes(x = beta.post.pn.0, y = beta.post.pn.1, color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(
    low = "lightskyblue",mid = "gold", high = "orangered", midpoint = 3) +
  labs(
    x = "alpha = 0",
    y = "alpha = 1",
    color = "s.e.",
    title = "posterior estimates") +
  theme_cowplot(font_size = 12)

# Compare the distributions of the estimated coefs.
x <- with(quantiles,c(q.pn.0,q.pn.1))
p3 <- ggplot(quantiles, aes(x = q.pn.0, y = q.pn.1)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "skyblue", linetype = "dashed") +
  coord_cartesian(xlim = range(x), ylim = range(x)) +
  labs(x = "alpha = 0",y = "alpha = 1",title = "quantiles") +
  theme_cowplot(font_size = 12)
plot_grid(p1,p2,p3,nrow = 1,ncol = 3)
```


```{r loglike.vs.alpha.point.normal}
ggplot(loglik %>% filter(alpha < 0.75), aes(x = alpha,y = loglik.point.normal)) +
  geom_point() +
  geom_line() +
  ggtitle("point normal: loglik vs. alpha") +
  theme_cowplot(font_size = 12)
```

For large $\alpha$ the point normal is not behaving very well. We estimate a large $\hat \pi_0$ and a large $\sigma_0$. Effects that the model is confident are non-zero experience virtually no shrinkage.


```{r alpha.v.sigma0}
library(latex2exp)
ggplot(params, aes(x=pi0.pn, y=sigma0.pn, color=alpha)) +
  geom_point() +
  geom_line() + 
  scale_color_gradient2(
    low = "lightskyblue",mid = "gold", high = "orangered", midpoint = 0.5) +
  labs(
    y = TeX('$\\sigma_0$'),
    x = TeX('$\\hat{\\pi}_0')
  )
```

```{r shirnkage.vs.responsibility}
p1 <- ggplot(pdat, aes(x=resp.pn.0, y=1 - (beta.hat - beta.post.pn.0) / beta.hat, color=se)) +
  geom_point() +
  scale_color_gradient2(
    low = "lightskyblue",mid = "gold", high = "orangered", midpoint = 3) +
  labs(
    x=TeX('p(\\beta != 0 | D)'),
    y='Prop shrinkage',
    title='Point normal alpha=0'
  )


p2 <- ggplot(pdat, aes(x=resp.pn.1, y=1 - (beta.hat - beta.post.pn.1)/beta.hat, color=se)) + 
  geom_point() + 
  scale_color_gradient2(
    low = "lightskyblue",mid = "gold", high = "orangered", midpoint = 3)+
  labs(
    x=TeX('p(\\beta != 0 | D)'),
    y='Prop shrinkage',
    title='Point normal alpha=1'
  )

p3 <- ggplot(pdat, aes(x=resp.pn.1, y=resp.pn.0, color=se)) +
  geom_point() +
  scale_color_gradient2(
    low = "lightskyblue",mid = "gold", high = "orangered", midpoint = 3)+
  labs(
    x=TeX('\\alpha = 1'),
    y=TeX('\\alpha = 0'),
    title=TeX('p(\\beta != 0 | D)')
  )  

cowplot::plot_grid(p1, p2, p3, ncol=3)
```


```{r se.quantile.hist}
se.quantile <- quantile(se, seq(0, 1, by=0.25))
bins <- cut(se, breaks=se.quantile, labels=paste0(seq(25, 100, by=25), '%'))

p1 <- pdat %>%
  mutate(se.bin = bins) %>%
  ggplot(aes(x=resp.pn.0, fill=se.bin)) +
  geom_histogram(alpha=0.5) +
  labs(
    x=TeX('p(\\beta != 0 | D)'),
    title=TeX('\\alpha = 0')
  )
p2 <- pdat %>%
  mutate(se.bin = bins) %>%
  ggplot(aes(x=resp.pn.1, fill=se.bin)) +
  geom_histogram(alpha=0.5) +
  labs(
    x=TeX('p(\\beta != 0 | D)'),
    title=TeX('\\alpha = 1')
  )
cowplot::plot_grid(p1, p2, ncol=2)
```


```{r}
ash.fits[[1]]$result %>% head()
```

```{r}
knitr::knit_exit()
```






```{r}
ggplot(pdat, aes(x=se, y=pi0.pn.1)) + geom_point()

ggplot(pdat, aes(x=resp.pn.1)) + geom_histogram()
ggplot(pdat, aes(x=resp.pn.0)) + geom_histogram()


```

```{r}
cor(quantiles)

pdat %>%
  select(contains('post')) %>%
  cor()

pdat %>% head
```

```{r ash.and.point.normal}
# Plot posterior estimates for best fit vs. worst fit.
pdat1 <- data.frame(b1 = b,
                    b2 = gsea.fits[[1]]$PosteriorMean,
                    se = se)
p1 <- ggplot(pdat2,aes(x = b1,y = b2,color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = 3) +
  labs(x = "beta hat",y = "beta post",color = "s.e.",
       title = "alpha=0 (Point normal)") +
  theme_cowplot(font_size = 12)

pdat2 <- data.frame(b1 = b,
                    b2 = gsea.fits[[n]]$PosteriorMean,
                    se = se)
p2 <- ggplot(pdat2, aes(x = b1,y = b2,color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = 3) +
  labs(x = "beta hat",y = "beta post",color = "s.e.",
       title = "alpha = 1 (Point normal)") +
  theme_cowplot(font_size = 12)


pdat3 <- data.frame(b1 = b,
                    b2 = fits[[1]]$result$PosteriorMean,
                    se = se)
p3 <- ggplot(pdat3,aes(x = b1,y = b2,color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = 3) +
  labs(x = "beta hat",y = "beta post",color = "s.e.",
       title = "alpha=0 (ASH)") +
  theme_cowplot(font_size = 12)

pdat4 <- data.frame(b1 = b,
                    b2 = fits[[n]]$result$PosteriorMean,
                    se = se)
p4 <- ggplot(pdat4,aes(x = b1,y = b2,color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = 3) +
  labs(x = "beta hat",y = "beta post",color = "s.e.",
       title = "alpha=1 (ASH)") +
  theme_cowplot(font_size = 12)

plot_grid(p1,p2,p3,p4, nrow = 2)
```




```{r ash.and.point.normal2}
# Compare the distributions of the estimated coefs.
pdat1 <-
  data.frame(q1 = quantile(gsea.fits[[1]]$PosteriorMean,seq(0.05,0.95,0.01)),
             q2 = quantile(fits[[1]]$result$PosteriorMean,seq(0.05,0.95,0.01)))
x <- with(pdat1,c(q1,q2))
p1 <- ggplot(pdat1,aes(x = q1,y = q2)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "skyblue",linetype = "dashed") +
  coord_cartesian(xlim = range(x),ylim = range(x)) +
  labs(x = "Point normal",y = "Ash",title = "quantiles, alpha=0") +
  theme_cowplot(font_size = 12)

pdat2 <-
  data.frame(q1 = quantile(gsea.fits[[n]]$PosteriorMean,seq(0.05,0.95,0.01)),
             q2 = quantile(fits[[n]]$result$PosteriorMean,seq(0.05,0.95,0.01)))
x <- with(pdat2,c(q1,q2))
p2 <- ggplot(pdat2,aes(x = q1,y = q2)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "skyblue",linetype = "dashed") +
  coord_cartesian(xlim = range(x),ylim = range(x)) +
  labs(x = "Point normal",y = "Ash",title = "quantiles, alpha=1") +
  theme_cowplot(font_size = 12)

pdat2 <-
  data.frame(q1 = quantile(gsea.fits[[1]]$PosteriorMean,seq(0.05,0.95,0.01)),
             q2 = quantile(fits[[n]]$result$PosteriorMean,seq(0.05,0.95,0.01)))
x <- with(pdat2,c(q1,q2))
p3 <- ggplot(pdat2,aes(x = q1,y = q2)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "skyblue",linetype = "dashed") +
  coord_cartesian(xlim = range(x),ylim = range(x)) +
  labs(x = "Point normal (alpha=0)",y = "Ash (alpha=1)",title = "quantiles") +
  theme_cowplot(font_size = 12)
plot_grid(p1,p2,p3, nrow = 1)
```


```{r}
# Plot posterior estimates for best fit vs. worst fit.
pdat1 <- data.frame(b1 = b,
                    b2 = gsea.fits[[1]]$PosteriorMean,
                    se = se)
p1 <- ggplot(pdat1,aes(y = b1 - b2, x = b1, color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = 3) +
  labs(y = "beta hat - beta_post", x = "beta post",color = "s.e.",
       title = "alpha=0 (Point normal)") +
  theme_cowplot(font_size = 12)

pdat1 <- data.frame(b1 = b,
                    b2 = gsea.fits[[n]]$PosteriorMean,
                    se = se)
p2 <- ggplot(pdat1,aes(y = b1 - b2, x = b1, color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = 3) +
  labs(y = "beta hat - beta_post", x = "beta post",color = "s.e.",
       title = "alpha=1 (Point normal)") +
  theme_cowplot(font_size = 12)

cowplot::plot_grid(p1, p2, nrow = 1)
```

```{r}
pdat1 <- data.frame(b1 = b,
                    b2 = gsea.fits[[1]]$PosteriorMean,
                    gamma = gsea.fits[[1]]$responsibilities,
                    se = se)

pdat1
p1 <- ggplot(pdat1,aes(y = b1 - b2, x = b1, color = se)) +
  geom_point() +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_color_gradient2(low = "lightskyblue",mid = "gold",high = "orangered",
                        midpoint = 3) +
  labs(y = "beta hat - beta_post", x = "beta post",color = "s.e.",
       title = "alpha=0 (Point normal)") +
  theme_cowplot(font_size = 12)
```


```{r}

```


```{r}
pis <- map(fits, ~pluck(.x, 'fitted_g', 'pi'))

map_int(pis, ~ length(.x))

do.call('cbind', pis) %>% dim
```



