---
title: "Baboon Diet Example"
author: "karltayeb"
date: "2022-03-16"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
params:
  cache_dir: 'cache/baboon_diet/'
  genesets: c('kegg', 'gomf', 'gobp')
  thresh: 1e-4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '#>',
  message=FALSE, warning=FALSE
)
```

```{r load.libraries}
library(tidyverse)
library(htmltools)
source('code/utils.R')
source('code/logistic_susie_vb.R')
source('code/logistic_susie_veb_boost.R')
source('code/load_gene_sets.R')
source('code/enrichment_pipeline.R')
```

## Introduction

Baboons were fed different diets for two years. Adipose, liver and muscle tissue were collected. Differential expression was conducted across the diet conditions.

## Setup
```{r load.and.standardize.data}
genesets <- load_gene_sets()

de.adipose <- read.table('data/wenhe_baboon_diet/DE_lrt_adipose.txt')
de.liver <- read.table('data/wenhe_baboon_diet/DE_lrt_liver.txt')
de.muscle <- read.table('data/wenhe_baboon_diet/DE_lrt_muscle.txt')

de <- bind_rows(
  rownames_to_column(de.adipose) %>% mutate(tissue='Adipose'), 
  rownames_to_column(de.liver) %>% mutate(tissue='Liver'),
  rownames_to_column(de.muscle) %>% mutate(tissue='Muscle')) %>%
  rename(gene=rowname)

hs <- org.Hs.eg.db::org.Hs.eg.db
gene_symbols <- unique(de$gene)
symbol2entrez <- AnnotationDbi::select(
  hs, keys=gene_symbols,
  columns=c('ENTREZID', 'SYMBOL'),
  keytype = 'SYMBOL')

add_names = function(l, n){
  names(l) <- n
  return(l)
}

data <- de %>%
  rename(SYMBOL=gene) %>%
  left_join(symbol2entrez, by='SYMBOL') %>%
  relocate(ENTREZID, .after=SYMBOL) %>%
  mutate(  # set default columns
    beta = logFC,
    se = 0,
    threshold.on = PValue
  ) %>%
  group_by(tissue) %>%
  group_map(~ .x, .keep = T) %>%
  add_names(map_chr(., ~pluck(.x, 'tissue')[1]))
```

### Fit 

```{r fit}
# fit logistic susie
do_logistic_susie_cached = function(data,
                                    db,
                                    thresh,
                                    prefix=''){
  res <- xfun::cache_rds({
    purrr::map_dfr(
      names(data),
      ~do_logistic_susie(.x, db, thresh, genesets, data))
  },
  dir = params$cache_dir,
  file=paste0(prefix, 'logistic_susie_', db, '_', thresh),
  hash = list(data, db, thresh, prefix))
}

params.genesets <- eval(parse(text=params$genesets))
params.thresh <- eval(parse(text=params$thresh))
fits <- map_dfr(params.genesets, ~do_logistic_susie_cached(data, .x, params.thresh))

# fit ora
do_ora_cached = function(data, db, thresh, prefix='', ...){
  res <- xfun::cache_rds({
    purrr::map_dfr(names(data), ~do_ora(.x, db, thresh, genesets, data))
  }, dir = params$cache_dir, file=paste0(prefix, 'ora_', db, '_', thresh), ...)
}
ora <- map_dfr(params.genesets, ~do_ora_cached(data, .x, params.thresh))
```

## Overview
### Threshold sensitivity

```{r threshold.sensitivity}
mean.gene.prop = function(l){
  purrr::map_dbl(3:10, ~get_y(l, 10^(-.x)) %>% mean())
}

thresh <- map_dbl(1:10, ~10**-.x)

.prop.ones = function(experiment){
  map_dbl(thresh, ~ prep_binary_data(
  genesets[['gobp']], data[[experiment]], thresh=.x)$y %>% mean())
}

prop.ones <- xfun::cache_rds({map_dfc(names(data), ~.prop.ones(.x))},
                             dir=params$cache_dir,
                             file='threshold_sensitivity')
colnames(prop.ones) <- names(data)
prop.ones <- prop.ones %>% mutate(thresh = thresh)

prop.ones %>%
  pivot_longer(one_of(names(data))) %>%
  group_by(name) %>%
  mutate(value = value) %>%
  ggplot(aes(x=factor(-log10(thresh)), y=value)) +
  geom_boxplot() +
  labs(
    y = 'Proportions of genes in gene list',
    title = 'Sensitivity to thresholding'
  )
```

```{r process.results, echo=FALSE, warning=FALSE}
# gene set summary
res <- fits %>%
  left_join(ora) %>%
  mutate(
    gs_summary = map(fit, get_gene_set_summary),
    res = map2(gs_summary, ora, ~ left_join(.x, .y)),
  ) %>%
  select(-c(fit, ora, gs_summary)) %>%
  unnest(res)

# credible set summary
res2 <- fits %>%
  left_join(ora) %>%
  mutate(res = map(fit, get_credible_set_summary)) %>%
  mutate(res = map2(res, ora, ~ left_join(.x, .y))) %>%
  select(-c(fit, ora)) %>%
  unnest(res)

# Setup HTML tables
split_tibble = function(tibble, col = 'col'){
  tibble %>% split(., .[, col])
}
descriptions <- map_dfr(genesets, ~pluck(.x, 'geneSet', 'geneSetDes'))
tbl <- res2 %>% 
  filter(active_cs) %>%
  left_join(descriptions)
tbl_split <- split_tibble(tbl, 'experiment')
html_tables <- map(tbl_split, ~split_tibble(.x, 'db'))
```


### Big volcano plot

Colors represent enrichment/depletion detected by Fishers exact test (Benjamini Hochberg corrected p-values < $0.05$). Gene sets that belong to a SuSiE credible set are circled.
```{r big.volcano.plot}
get_ora_enrichments = function(tbl){
   tbl %>% mutate(
    padj = p.adjust(pFishersExact),
    result = case_when(
      padj < 0.05 & oddsRatio < 1 ~ 'depleted',
      padj < 0.05 & oddsRatio > 1 ~ 'enriched',
      TRUE ~ 'not significant'
    )
  )
}

# plot all enrichments, highlight gene sets in credible set
csdat <- res2 %>% 
  filter(in_cs, active_cs)

res %>% 
  group_by(experiment, db) %>%
  get_ora_enrichments %>%
  ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
  geom_point() +
  geom_point(
    csdat, 
    mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact)),
    color='black', pch=21, size=5) +
  scale_color_manual(values = c('depleted' = 'coral',
                                'enriched' = 'dodgerblue',
                                'not significant' = 'grey')) +
  facet_wrap(vars(db))
```

## Enrichment results

### Glossary

* `alpha` is the posterior probability of SuSiE including this gene set **in this component** which is different from PIP (probability of SuSiE including this gene set in ANY component)
* `beta` posterior mean/standard error of posterior mean for effect size. Standard errors are likely too small. 
* `oddsRatio, pHypergeometric, pFishersExact` construct a contingency table (gene list membersip) x (gene set membership), estimate the `oddsRatio` gives the odds of being in the gene list conditional on being in the gene set / odds of being in the gene list conditional on NOT being in the gene set. `pHypergeometric` and `pFishersExact` are pvalues from 1 and 2 sided test respectively. 
```{r experiment.level.results, results='asis'}
experiments <- unique(res$experiment)

do.experiment.volcano = function(this_experiment){
  res %>% 
    filter(experiment == this_experiment) %>%
    group_by(db) %>%
    get_ora_enrichments %>%
    ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
    geom_point() +
    geom_point(
      csdat %>% filter(experiment == this_experiment), 
      mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact)),
      color='black', pch=21, size=5) +
    scale_color_manual(values = c('depleted' = 'coral',
                                  'enriched' = 'dodgerblue',
                                  'not significant' = 'grey')) +
    facet_wrap(vars(db)) +
    labs(title = this_experiment)
}

for (i in 1:length(experiments)){
  this_experiment <- experiments[i]
  cat("\n") 
  cat("###", this_experiment, "\n") # Create second level headings with the names.
  do.experiment.volcano(this_experiment) %>% print()
  cat("\n\n") 

  for(db in names(html_tables[[this_experiment]])){
    cat("####", db, "\n") # Create second level headings with the names.
    to_print <- html_tables[[this_experiment]][[db]] %>% distinct()
    to_print %>% report_susie_credible_sets() %>% htmltools::HTML() %>% print()
    cat("\n")
  }}
```

