---
title: "Baboon Diet Example"
author: "karltayeb"
date: "2022-03-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Baboons were fed different diets for two years. Adipose, liver and muscle tissue were collected. Differential expression was conducted across the diet conditions.

```{r message=FALSE, results='hide'}
library(GSEABenchmarkeR)
library(EnrichmentBrowser)
library(tidyverse)
library(susieR)
library(htmltools)
library(DT)
library(kableExtra)

source('code/load_gene_sets.R')
source('code/utils.R')
source('code/logistic_susie_vb.R')
source('code/logistic_susie_veb_boost.R')
source('code/latent_logistic_susie.R')
```

## Setup

### Load Gene Sets
`loadGeneSetX` uniformly formats gene sets and generates the $X$ matrix
We can source any gene set from `WebGestaltR::listGeneSet()`

```{r geneset.setup, message=FALSE, results='hide'}
gs_list <- WebGestaltR::listGeneSet()
gobp <- loadGeneSetX('geneontology_Biological_Process', min.size=50)  # just huge number of gene sets
gobp_nr <- loadGeneSetX('geneontology_Biological_Process_noRedundant', min.size=1)
gomf <- loadGeneSetX('geneontology_Molecular_Function', min.size=1)
kegg <- loadGeneSetX('pathway_KEGG', min.size=1)
reactome <- loadGeneSetX('pathway_Reactome', min.size=1)
wikipathway_cancer <- loadGeneSetX('pathway_Wikipathway_cancer', min.size=1)
wikipathway <- loadGeneSetX('pathway_Wikipathway', min.size=1)

genesets <- list(
  gobp=gobp,
  gobp_nr=gobp_nr,
  gomf=gomf,
  kegg=kegg,
  reactome=reactome,
  wikipathway_cancer=wikipathway_cancer,
  wikipathway=wikipathway
)
```

```{r data.setup}
adipose <- read.table('data/wenhe_baboon_diet/DE_lrt_adipose.txt')
muscle <- read.table('data/wenhe_baboon_diet/DE_lrt_muscle.txt')
liver <- read.table('data/wenhe_baboon_diet/DE_lrt_liver.txt')
deseq <- list(Adipose=adipose, Muscle = muscle, Liver = liver)

convert_labels <- function(y, from='SYMBOL', to='ENTREZID'){
  hs <- org.Hs.eg.db::org.Hs.eg.db
  gene_symbols <- names(y)
  symbol2entrez <- AnnotationDbi::select(hs, keys=gene_symbols, columns=c(to, from), keytype = from)
  symbol2entrez <- symbol2entrez[!duplicated(symbol2entrez[[from]]),]
  symbol2entrez <- symbol2entrez[!is.na(symbol2entrez[[to]]),]
  symbol2entrez <- symbol2entrez[!is.na(symbol2entrez[[from]]),]
  rownames(symbol2entrez) <- symbol2entrez[[from]]
  ysub <- y[names(y) %in% symbol2entrez[[from]]]
  names(ysub) <- symbol2entrez[names(ysub),][[to]]
  return(ysub)
}

par(mfrow=c(1,1))
deseq$Adipose %>% .$PValue %>% hist(main='Adipose p-values')
```


### Fitting and reporting helpers

```{r driver}
#' take gene level results
#' put into standard format
#' call the value we threshold on 'y'
#' convert gene labels to ENTREIZID
clean_gene_list = function(dat, thresh_col='padj', from='SYMBOL'){
  thresh_col = sym(thresh_col)
  dat %>%
    as.data.frame %>%
    rownames_to_column('gene') %>%
    dplyr::select(gene, !!thresh_col) %>%
    filter(!is.na(!!thresh_col)) %>%
    mutate(y = !!thresh_col) %>%
    select(gene, y) %>%
    tibble2namedlist %>%
    convert_labels(from)
}

#' modify this for different data sources
#' makes a binary list from table like data
proc_y = function(data, thresh){
  data %>%
    clean_gene_list('PValue') %>%
    {
      y <- as.integer(. < thresh)
      names(y) <- names(.)
      y
    }
}


#' fit logistic susie, and hypergeometric test
logistic_susie_driver = function(db, celltype, thresh){
  gs <- genesets[[db]]
  data <- deseq[[celltype]]

  # set up binary y
  y <- proc_y(data, thresh)
  
  u <- process_input(gs$X, y)  # subset to common genes
  vb.fit <- logistic.susie(  # fit model
    u$X, u$y, L=10, init.intercept = 0, verbose=1, maxit=100)

  #' hypergeometric test
  ora <- tibble(
    geneSet = colnames(u$X),
    geneListSize = sum(u$y),
    geneSetSize = colSums(u$X),
    overlap = (u$y %*% u$X)[1,],
    nGenes = length(u$y),
    propInList = overlap / geneListSize,
    propInSet = overlap / geneSetSize,
    oddsRatio = (overlap / (geneListSize - overlap)) / (
      (geneSetSize - overlap) / (nGenes - geneSetSize + overlap)),
    pValueHypergeometric = phyper(
      overlap-1, geneListSize, nGenes, geneSetSize, lower.tail= FALSE),
    db = db,
    celltype = celltype,
    thresh = thresh
  ) %>% 
  left_join(gs$geneSet$geneSetDes)

  return(list(
    fit = vb.fit,
    ora = ora,
    db = db, celltype = celltype, thresh = thresh))
}

credible_set_summary = function(res){
  gs <- genesets[[res$db]]
  data <- deseq[[res$celltype]]

  # set up binary y
  y <- proc_y(data, thresh)
  
  u <- process_input(gs$X, y)  # subset to common genes

  #' report top 50 elements in cs
  credible.set.summary <- t(res$fit$alpha) %>%
    data.frame() %>%
    rownames_to_column(var='geneSet') %>%
    rename_with(~str_replace(., 'X', 'L')) %>%
    pivot_longer(starts_with('L'), names_to='component', values_to = 'alpha') %>%
    arrange(component, desc(alpha)) %>%
    dplyr::group_by(component) %>%
    filter(row_number() < 50) %>%
    mutate(cumalpha = c(0, head(cumsum(alpha), -1))) %>%
    mutate(in_cs = cumalpha < 0.95) %>%
    mutate(active_cs = component %in% names(res$fit$sets$cs)) %>%
    left_join(res$ora) %>%
    left_join(gs$geneSet$geneSetDes)

  #' map each gene set to the component with top alpha
  #' report pip
  gene.set.summary <- res$fit$pip %>% 
    as_tibble(rownames='geneSet') %>%
    rename(pip=value) %>%
    left_join(res$ora) %>%
    left_join(gs$geneSet$geneSetDes)

  return(credible.set.summary)
}

gene_set_summary = function(res){
  gs <- genesets[[res$db]]
  #' map each gene set to the component with top alpha
  #' report pip
  res$fit$pip %>% 
    as_tibble(rownames='geneSet') %>%
    rename(pip=value) %>%
    left_join(res$ora) %>%
    left_join(gs$geneSet$geneSetDes)
}
```


### Example usage

```{r}
db = 'gobp'
celltype = 'Adipose'
thresh = 1e-4
  
res <- logistic_susie_driver(db, celltype, thresh)
credible_set_summary(res) %>% head()
gene_set_summary(res) %>% head()
```

## Fit logistic SuSiE 

For each celltype, we fit logistic SuSiE using multiple gene set sources at various threshold of `padj`.

```{r run.all, results='hide', message=FALSE}
celltypes <- names(deseq)
pthresh <- c(1e-2) # c(0.1, 0.01, 0.001, 0.0001, 0.00001, 0.000001)
db_name <- names(genesets)
crossed <- cross3(db_name, celltypes, pthresh)

res <- xfun::cache_rds({
  res <- purrr::map(crossed, purrr::lift_dl(logistic_susie_driver))
  for (i in 1:length(res)){  # save some space
    res[[i]]$fit$dat <- NULL
  }
  res
  }
)

u <- map(res, credible_set_summary)
cs.summary <- do.call('rbind', u)

u <- map(res, gene_set_summary)
geneset.summary <- do.call('rbind', u)
```

### Summary functions

Just a few functions to help streamline looking at output

```{r report.result.functions}
library(kableExtra)

#' takes a tibble
#' organize by database and component
#' report credible set, descriptions, pips, and hypergeometric pvalue
#' in one row, with cs ordered by pip
db_component_kable = function(tbl, min_alpha=0, max_sets=10){
  tbl_filtered <-
    tbl %>%
    group_by(celltype, db, component) %>%
    arrange(celltype, db, component, desc(alpha)) %>%
    filter(alpha > min_alpha, row_number() <= max_sets)

  tbl_filtered %>%
    select(geneSet, description, alpha, pValueHypergeometric) %>%
    dplyr::mutate_if(is.numeric, funs(as.character(signif(., 3)))) %>%
    kbl() %>%
    column_spec(c(4, 5, 6), color=ifelse(tbl_filtered$in_cs, 'green', 'red')) %>%
    collapse_rows(columns = c(1, 2), valign = 'top') %>%
    kableExtra::kable_styling()
}
```
  

## Results

Gene set results for each tissue, across multiple gene set sources.

We show the top 10 gene sets per credible set, green indicates that the gene set is in the 95% credible set. Red indicates it's not in the 95% credible set, but still ranked highly for SuSiE that component, to help get a sense of what's getting lumped in with the enrichments reported by SuSiE.


```{r, results="asis", message=FALSE}
gen.table = function(ct, g){
  cs.summary %>%
        filter(celltype == ct, db == g, active_cs) %>%
        db_component_kable() %>%
        htmltools::HTML()
}
possibly.gen.table = possibly(gen.table, otherwise="nothing to report...")

for(ct in celltypes){
  cat("\n") 
  cat("##", ct, "\n") # Create second level headings with the names.
  
  for(g in names(genesets)){
    cat("###", g, "\n") # Create second level headings with the names.
    possibly.gen.table(ct, g) %>% print()
    cat("\n")
  }
}
```

```{r}
knitr::knit_exit()
```

