---
title: "Deng et al. Logistic SuSiE GSEA"
author: "karltayeb"
date: "2022-04-03"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
params:
  cache_dir: 'cache/deng/'
  genesets: c('gobp_nr')
  thresh: 1e-4
  rerun: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '',
  message=FALSE, warning=FALSE
)
```

## Overview

We report gene set enrichment analysis of the topics from [Jason's NMF and SNMF](https://willwerscheid.github.io/scFLASH/deng_nn.html) of the Deng et al. data set. 

For each topic we construct a binary gene list by thresholding on local false sign rates output from `flash`. Here, I pick $lfsr <- 1e-5$.


```{r load.libraries}
library(tidyverse)
library(htmltools)
source('code/utils.R')
```


```{r load.data}
source('code/load_gene_sets.R')
source('code/load_data.R')
data <- load_deng_topics()
genesets <- load_gene_sets()
```


```{r fit}
source('code/logistic_susie_vb.R')
source('code/logistic_susie_veb_boost.R')
source('code/enrichment_pipeline.R')

# fix params
params.genesets <- eval(parse(text=params$genesets))
params.thresh <- eval(parse(text=params$thresh))
params.rerun <- eval(parse(text=params$rerun))

do_logistic_susie_cached = function(data, db, thresh, .sign=c(1, -1), prefix=''){
  res <- xfun::cache_rds({
    purrr::map_dfr(
      names(data),
      ~do_logistic_susie(.x, db, thresh, genesets, data, .sign=.sign))},
    dir = params$cache_dir,
    file=paste0(prefix, 'logistic_susie_', db, '_', thresh),
    hash = list(data, db, thresh, prefix),
    rerun = params.rerun)
}

fits <- map_dfr(params.genesets, ~do_logistic_susie_cached(data, .x, params.thresh))

snmf.data <- data[grep('^SNMF*', names(data))]
positive.fits <- map_dfr(
  params.genesets,
  ~do_logistic_susie_cached(snmf.data, .x, params.thresh, c(1), 'positive_'))
negative.fits <- map_dfr(
  params.genesets,
  ~do_logistic_susie_cached(snmf.data, .x, params.thresh, c(-1), 'negative_'))

do_ora_cached = function(data, db, thresh, .sign=c(1, -1), prefix=''){
  res <- xfun::cache_rds({
    purrr::map_dfr(names(data), ~do_ora(.x, db, thresh, genesets, data, .sign=.sign))
  },
  dir = params$cache_dir,
  file=paste0(prefix, 'ora_', db, '_', thresh),
  rerun=params.rerun)
}

ora <- map_dfr(params.genesets, ~do_ora_cached(data, .x, params.thresh))
positive.ora <- map_dfr(
  params.genesets,
  ~do_ora_cached(snmf.data, .x, params.thresh, c(1), 'positive_'))
negative.ora <- map_dfr(
  params.genesets,
  ~do_ora_cached(snmf.data, .x, params.thresh, c(-1), 'negative_'))
```

```{r summarise.fits}
# take credible set summary, return "best" row for each gene set
get_cs_summary_condensed = function(fit){
  fit %>%
    get_credible_set_summary() %>%
    group_by(geneSet) %>%
    arrange(desc(alpha)) %>%
    filter(row_number() == 1)
}
# generate table for making gene-set plots
get_plot_tbl = function(fits, ora){
  res <- fits %>%
    left_join(ora) %>%
    mutate(
      gs_summary = map(fit, get_gene_set_summary),
      cs_summary = map(fit, get_cs_summary_condensed),
      res = map2(gs_summary, cs_summary, ~ left_join(.x, .y, by='geneSet')),
      res = map2(res, ora, ~ possibly(left_join, NULL)(.x, .y))
    ) %>%
    select(-c(fit, susie.args, ora, gs_summary, cs_summary)) %>%
    unnest(res)
  return(res)
}

# split tibble into a list using 'col'
split_tibble = function(tibble, col = 'col'){
  tibble %>% split(., .[, col])
}
  
# Get summary of credible sets with gene set descriptions
get_table_tbl = function(fits, ora){
  res2 <- fits %>%
    left_join(ora) %>%
    mutate(res = map(fit, get_credible_set_summary)) %>%
    mutate(res = map2(res, ora, ~ left_join(.x, .y))) %>%
    select(-c(fit, ora)) %>%
    unnest(res)

  descriptions <- map_dfr(genesets, ~pluck(.x, 'geneSet', 'geneSetDes'))
  tbl <- res2 %>% 
    filter(active_cs) %>%
    left_join(descriptions)
  tbl_split <- split_tibble(tbl, 'experiment')
  html_tables <- map(tbl_split, ~split_tibble(.x, 'db'))
  return(html_tables)
}
```


```{r}
get_ora_enrichments = function(tbl){
   tbl %>% mutate(
    padj = p.adjust(pFishersExact),
    result = case_when(
      padj < 0.05 & oddsRatio < 1 ~ 'depleted',
      padj < 0.05 & oddsRatio > 1 ~ 'enriched',
      TRUE ~ 'not significant'
    )
  )
}

do.volcano = function(res){
  res %>% 
    get_ora_enrichments %>%
    ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
    geom_point() +
    geom_point(
      res %>% filter(in_cs, active_cs), 
      mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact)),
      color='black', pch=21, size=5) +
    scale_color_manual(values = c('depleted' = 'coral',
                                  'enriched' = 'dodgerblue',
                                  'not significant' = 'grey'))
}
```

## NMF 

### Threshold sensitivity

```{r threshold.sensitivity}
thresh <- map_dbl(1:20, ~10**-.x)

.prop.ones = function(topic){
  map_dbl(thresh, ~ prep_binary_data(
  genesets[['gobp']], data[[topic]], thresh=.x)$y %>% mean())
}

prop.ones <- map_dfc(names(data[grep('^NMF*', names(data))]), ~.prop.ones(.x))

new.names = paste0('Topic', str_pad(2:20, 2, pad='0'))
colnames(prop.ones) <- new.names
prop.ones <- prop.ones %>% mutate(thresh = thresh)

prop.ones %>%
  pivot_longer(one_of(new.names)) %>%
  filter(thresh <= 1e-3) %>%
  group_by(name) %>%
  mutate(value = value / (sort(value, decreasing = T)[1])) %>%
  ggplot(aes(x=-log10(thresh), y=value, color=name)) +
  geom_point() + geom_line() +
  labs(
    y = 'prop.ones relative to 1e-3 threshold',
    title = 'Sensitivity of lfsr to threshold'
  )
```


### Glossary

* `alpha` is the posterior probability of SuSiE including this gene set **in this component** which is different from PIP (probability of SuSiE including this gene set in ANY component)
* `beta` posterior mean/standard error of posterior mean for effect size. Standard errors are likely too small. 
* `oddsRatio, pHypergeometric, pFishersExact` construct a contingency table (gene list membersip) x (gene set membership), estimate the `oddsRatio` gives the odds of being in the gene list conditional on being in the gene set / odds of being in the gene list conditional on NOT being in the gene set. `pHypergeometric` and `pFishersExact` are pvalues from 1 and 2 sided test respectively. 

```{r nmf.results, results='asis'}
res <- fits %>%
  filter(str_detect(experiment, '^NMF*')) %>%
  get_plot_tbl(., ora)

html_tables <- fits %>%
  filter(str_detect(experiment, '^NMF*')) %>%
  get_table_tbl(., ora)

experiments <- unique(res$experiment)

for (i in 1:length(experiments)){
  this_experiment <- experiments[i]
  cat("\n") 
  cat("###", this_experiment, "\n") # Create second level headings with the names.
  
  # print volcano plot
  p <- do.volcano(res %>% filter(experiment == this_experiment)) +
    labs(title=this_experiment)
  print(p)

  cat("\n\n") 

  # print table
  for(db in names(html_tables[[this_experiment]])){
    cat("####", db, "\n") # Create second level headings with the names.
    to_print <- html_tables[[this_experiment]][[db]] %>% distinct()
    to_print %>% report_susie_credible_sets() %>% htmltools::HTML() %>% print()
    cat("\n")
  }}
```


## SNMF 

```{r snmf.results, results='asis'}
res <- fits %>%
  filter(str_detect(experiment, '^SNMF*')) %>%
  get_plot_tbl(., ora)

html_tables <- fits %>%
  filter(str_detect(experiment, '^SNMF*')) %>%
  get_table_tbl(., ora)

experiments <- unique(res$experiment)
for (i in 1:length(experiments)){
  this_experiment <- experiments[i]
  cat("\n") 
  cat("###", this_experiment, "\n") # Create second level headings with the names.
  
  # print volcano plot
  p <- do.volcano(res %>% filter(experiment == this_experiment)) +
    labs(title=this_experiment)
  print(p)

  cat("\n\n") 

  # print table
  for(db in names(html_tables[[this_experiment]])){
    cat("####", db, "\n") # Create second level headings with the names.
    to_print <- html_tables[[this_experiment]][[db]] %>% distinct()
    to_print %>% report_susie_credible_sets() %>% htmltools::HTML() %>% print()
    cat("\n")
  }}
```


## SNMF - Positive loadings

```{r positive.snmf.results, results='asis'}
res <- positive.fits %>%
  filter(str_detect(experiment, '^SNMF*')) %>%
  get_plot_tbl(., positive.ora)

html_tables <- positive.fits %>%
  filter(str_detect(experiment, '^SNMF*')) %>%
  get_table_tbl(., positive.ora)

experiments <- unique(res$experiment)
for (i in 1:length(experiments)){
  this_experiment <- experiments[i]
  cat("\n") 
  cat("###", this_experiment, "\n") # Create second level headings with the names.
  
  # print volcano plot
  p <- do.volcano(res %>% filter(experiment == this_experiment)) +
    labs(title=this_experiment)
  print(p)

  cat("\n\n") 

  # print table
  for(db in names(html_tables[[this_experiment]])){
    cat("####", db, "\n") # Create second level headings with the names.
    to_print <- html_tables[[this_experiment]][[db]] %>% distinct()
    to_print %>% report_susie_credible_sets() %>% htmltools::HTML() %>% print()
    cat("\n")
  }}
```

## SNMF - Negative loadings

```{r negative.snmf.results, results='asis'}
res <- negative.fits %>%
  filter(str_detect(experiment, '^SNMF*')) %>%
  get_plot_tbl(., negative.ora)

html_tables <- negative.fits %>%
  filter(str_detect(experiment, '^SNMF*')) %>%
  get_table_tbl(., negative.ora)

experiments <- unique(res$experiment)
for (i in 1:length(experiments)){
  this_experiment <- experiments[i]
  cat("\n") 
  cat("###", this_experiment, "\n") # Create second level headings with the names.
  
  # print volcano plot
  p <- do.volcano(res %>% filter(experiment == this_experiment)) +
    labs(title=this_experiment)
  print(p)

  cat("\n\n") 

  # print table
  for(db in names(html_tables[[this_experiment]])){
    cat("####", db, "\n") # Create second level headings with the names.
    to_print <- html_tables[[this_experiment]][[db]] %>% distinct()
    to_print %>% report_susie_credible_sets() %>% htmltools::HTML() %>% print()
    cat("\n")
  }}
```

```{r}
knitr::knit_exit()
```
