---
title: "Logistic SuSiE Report Template"
author: "karltayeb"
date: "2022-04-11"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
params:
  cache_dir: 'cache/example_pbmc/'
  prefix: 'scpbmcdeseq'
  genesets: c('all_msigdb', 'all_go', 'pathways')
  thresh: 1e-4
  rerun: FALSE  # set to true to ignore cache
---

## Introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '#>',
  message=FALSE, warning=FALSE
)
```

```{r load.libraries, include=FALSE}
library(tidyverse)
library(htmltools)
devtools::load_all('~/R/gseasusie/') #library(gseasusie) 
library(tictoc)
source('code/utils.R')

params.cache_dir <- params$cache_dir
params.rerun <- eval(parse(text=params$rerun))
cache_rds <- partial(xfun::cache_rds, dir=params.cache_dir, rerun=params.rerun)
```

```{r load.data, include=FALSE}
source('code/load_data.R')
params.genesets <- eval(parse(text=params$genesets))
genesets <- gseasusie::load_gene_sets(c('c2', params.genesets))
data <- load_sc_pbmc_deseq2()
```


```{r}
run_enrichment = function(prefix, db, experiment, thresh){
  message(paste0('Fitting logistic susie...',
    '\n\tTitle = ', prefix,
    '\n\tExperiment = ', experiment,
    '\n\tDatabase = ', db,
    '\n\tthresh = ', thresh))
  
  res <- cache_rds({
  bin.data <- gseasusie::prep_binary_data(genesets[[db]], data[[experiment]], thresh)
  X <- bin.data$X
  y <- bin.data$y
  
  logistic.fit <- gseasusie::fit_logistic_susie_veb_boost(X, y, L=20)
  linear.fit <- susieR::susie(X, y)
  ora <- gseasusie::fit_ora(X, y)
  marginal_regression <- gseasusie::fit_marginal_regression_jax(X, y)
  residual_regression <- gseasusie::fit_residual_regression_jax(X, y, logistic.fit)

  res <- tibble(
    db = db,
    experiment = experiment,
    thresh = thresh,
    fit=list(logistic.fit),
    linear.fit = list(linear.fit),
    ora=list(ora),
    marginal_reg = list(marginal_regression),
    residual_reg = list(residual_regression)
  )
  res
  },  file=paste0(prefix, '_', experiment, '_', db, '_', thresh))
  return(res)
}

a <- run_enrichment('scpbmcdeseq', 'c2', 'CD14+ Monocyte', 1e-4)

summarise_enrichment = function(fit, ora, marginal_reg, residual_reg, geneSetDes){
  # put everything into one table
  gssum <- get_gene_set_summary(fit)
  cssum <- get_cs_summary_condensed(fit)
  
  reg <- left_join(
    marginal_reg,
    residual_reg,
    by="geneSet", suffix=c('_marginal', '_residual'))

  gene_set_summary <- 
    gssum %>%
    dplyr::left_join(cssum) %>%
    dplyr::left_join(ora) %>%
    left_join(reg) %>%
    left_join(geneSetDes)
  return(gene_set_summary)
}

b <- a %>% rowwise() %>% mutate(summary = list(summarise_enrichment(
  fit, ora, marginal_regression, residual_regression, genesets[[db]]$geneSetDes)))

b$summary[[1]] %>% filter(in_cs, active_cs)

gssum <- get_gene_set_summary(a$fit[[1]])
cssum <- get_cs_summary_condensed(a$fit[[1]])

reg <- left_join(
  a$marginal_reg[[1]],
  a$residual_reg[[1]],
  by="geneSet", suffix=c('_marginal', '_residual'))
  
geneSetDes <- genesets$c2$geneSetDes
gene_set_summary <- 
  gssum %>%
  dplyr::left_join(cssum) %>%
  dplyr::left_join(ora) %>%
  left_join(reg) %>%
  left_join(geneSetDes)

summarise_enrichment(a$fit[[1]], a$ora[[1]], a$marginal_reg[[1]], a$residual_reg[[1]], genesets$c2$geneSetDes)

a$residual_reg[[1]]

run_enrichment('scpbmcdeseq', 'all_go', 'CD14+ Monocyte', 1e-4)
run_enrichment('scpbmcdeseq', 'all_msigdb', 'CD14+ Monocyte', 1e-4)
run_enrichment('scpbmcdeseq', 'pathways', 'CD14+ Monocyte', 1e-4)


fit <- a$fit[[1]];
ora <- a$ora[[1]];
marginal_regression <- a$marginal_reg[[1]]
residual_regression <- a$residual_reg[[1]]


geneSetDes <- genesets$c2$geneSetDes

# put everything into one table
reg <- left_join(
  marginal_regression,
  residual_regression,
  by="geneSet", suffix=c('_marginal', '_residual'))

res <- get_gene_set_summary(fit) %>%
  dplyr::left_join(get_credible_set_summary(fit)) %>%
  dplyr::left_join(ora) %>%
  left_join(reg) %>%
  left_join(geneSetDes)

genesets$c2$geneSetDes


enrichment_volcano2 = function(res){
  res %>% 
    label_sig_enrichments() %>%
    ggplot2::ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
    ggplot2::geom_point() +
    ggplot2::geom_point(
      res %>% filter(in_cs, active_cs), 
      mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact), fill=component),
      color='black', pch=21, size=5) +
    ggplot2::scale_color_manual(values = c('depleted' = 'coral',
                                           'enriched' = 'dodgerblue',
                                           'not significant' = 'grey'))
}

residual_enrichment_histogram2 = function(res){
  res %>%
    dplyr::select(geneSet, pval_marginal, pval_residual) %>%
    tidyr::pivot_longer(dplyr::starts_with('pval'), values_to = 'pval') %>%
    ggplot2::ggplot(aes(x=pval)) +
      ggplot2::geom_histogram() + ggplot2::facet_wrap(vars(name))
}

```

```{r}
params.genesets <- eval(parse(text=params$genesets))
params.thresh <- eval(parse(text=params$thresh))

meta <- crossing(
  prefix = params$prefix,
  db=params.genesets, 
  experiment=names(data),
  thresh = params.thresh
)

meta %>% rowwise() %>% mutate(args = list(.))
meta2 <- split(as.data.frame(meta), seq(nrow(meta))) %>% map(as.list)
```


```{r}
library(furrr)
future::plan(multisession, workers = 6)

# take a list of kwargs and unpack it into run_enrichment
exec_run_enrichment = function(args) exec(run_enrichment, !!!args)
furrr::future_map_dfr(meta2, exec_run_enrichment)
```


```{r fit, include=FALSE}
params.genesets <- eval(parse(text=params$genesets))
params.thresh <- eval(parse(text=params$thresh))

# fit logistic susie
do_logistic_susie_cached = function(data,
                                    db,
                                    thresh,
                                    prefix=''){
  res <- cache_rds({
    purrr::map_dfr(
      names(data),
      ~do_logistic_susie_veb_boost(.x, db, thresh, genesets, data))
  },
  file=paste0(prefix, 'logistic_susie_', db, '_', thresh),
  hash = list(data, db, thresh, prefix))
}
fits <- map_dfr(params.genesets, ~do_logistic_susie_cached(data, .x, params.thresh))

# fit ora
do_ora_cached = function(data, db, thresh, prefix='', ...){
  res <- cache_rds({
    purrr::map_dfr(
      names(data),
      ~do_ora(.x, db, thresh, genesets, data))
  },
  file=paste0(prefix, 'ora_', db, '_', thresh),
  hash = list(data, db, thresh, prefix))
}
ora <- map_dfr(params.genesets, ~do_ora_cached(data, .x, params.thresh))
```


```{r residual.regression}
susie_coef = function(fit){
  return(colSums((fit$alpha * fit$mu)))
}

susie_pred = function(fit){
  return((fit$intercept + fit$dat$X %*% susie_coef(fit))[, 1])
}

residual_regression = function(y, x, o){
  X <- matrix(c(rep(1, length(x)), x), ncol = 2)
  summary(fastglm::fastglm(x=X, y=y, offset=o, family='binomial'))$coef %>% tail(1)
}

do_residual_regression = function(fit){
  pred <- susie_pred(fit)
  p <- dim(fit$dat$X)[2]

  message('performing logistic regression on residuals')
  tictoc::tic()
  residual_res <- map(1:p, ~residual_regression(fit$dat$y, fit$dat$X[, .x], pred))
  tictoc::toc()
  residual_res <- do.call(rbind, residual_res)
  residual_res <- as_tibble(residual_res) %>%  mutate(geneSet = colnames(fit$dat$X))
  colnames(residual_res) <- c('residual_beta', 'residual_se', 'residual_z', 'residual_p', 'geneSet')
  return(residual_res)
}

fits <- cache_rds({
  message('computing residual regressions')
  fits %>% rowwise() %>% mutate(residual_regression = list(do_residual_regression(fit))) %>% ungroup()
}, file='residual_regression')
```


## Overview
### Threshold sensitivity

```{r threshold.sensitivity}
ttt <- map_dbl(1:20, ~10**-.x)

.prop.ones = function(experiment){
  map_dbl(ttt, ~ prep_binary_data(
  genesets[['gobp_nr']], data[[experiment]], thresh=.x)$y %>% mean())
}

prop.ones <- cache_rds({map_dfc(names(data), ~.prop.ones(.x))}, file='threshold_sensitivity')
colnames(prop.ones) <- names(data)
prop.ones <- prop.ones %>% mutate(thresh = ttt)

prop.ones %>%
  pivot_longer(one_of(names(data))) %>%
  group_by(name) %>%
  mutate(value = value) %>%
  ggplot(aes(x=factor(-log10(thresh)), y=value)) +
  geom_boxplot() +
  labs(
    y = 'Proportions of genes in gene list',
    title = 'Sensitivity to thresholding'
  )
```

```{r process.results, echo=FALSE, warning=FALSE}
# gene set summary
res <- fits %>%
  left_join(ora) %>%
  mutate(
    gs_summary = map(fit, gseasusie::get_gene_set_summary),
    res = map2(gs_summary, ora, ~ left_join(.x, .y)),
    res = map2(res, residual_regression, ~ left_join(.x, .y)),
  ) %>%
  select(-c(fit, residual_regression, ora, gs_summary)) %>%
  unnest(res)

# credible set summary
res2 <- fits %>%
  left_join(ora) %>%
  mutate(res = map(fit, gseasusie::get_credible_set_summary)) %>%
  mutate(res = map2(res, ora, ~ left_join(.x, .y))) %>%
  mutate(res = map2(res, residual_regression, ~ left_join(.x, .y))) %>%
  select(-c(fit, ora)) %>%
  unnest(res)

# Setup HTML tables
split_tibble = function(tibble, col = 'col'){
  tibble %>% split(., .[, col])
}

descriptions <- map_dfr(genesets, ~pluck(.x, 'geneSet', 'geneSetDes'))
tbl <- res2 %>% 
  filter(active_cs, in_cs)
tbl_split <- split_tibble(tbl, 'experiment')
html_tables <- map(tbl_split, ~split_tibble(.x, 'db'))
# nested list like html_tables[[experiment]][[db]]

```


### Big volcano plot

Colors represent enrichment/depletion detected by Fishers exact test (Benjamini Hochberg corrected p-values < $0.05$). Gene sets that belong to a SuSiE credible set are circled.

```{r big.volcano.plot}
get_ora_enrichments = function(tbl){
   tbl %>% mutate(
    padj = p.adjust(pFishersExact),
    result = case_when(
      padj < 0.05 & oddsRatio < 1 ~ 'depleted',
      padj < 0.05 & oddsRatio > 1 ~ 'enriched',
      TRUE ~ 'not significant'
    )
  )
}

get_residual_enrichments = function(tbl){
   tbl %>% mutate(
    padj = p.adjust(residual_p),
    result = case_when(
      padj < 0.05 & residual_beta < 1 ~ 'depleted',
      padj < 0.05 & residual_beta > 1 ~ 'enriched',
      TRUE ~ 'not significant'
    )
  )
}

# plot all enrichments, highlight gene sets in credible set
csdat <- res2 %>% 
  filter(in_cs, active_cs)

res %>% 
  group_by(experiment, db) %>%
  get_ora_enrichments %>%
  ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
  geom_point() +
  geom_point(
    csdat, 
    mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact), fill=component),
    color='black', pch=21, size=5) +
  scale_color_manual(values = c('depleted' = 'coral',
                                'enriched' = 'dodgerblue',
                                'not significant' = 'grey')) +
  facet_wrap(vars(db))


res %>%
  group_by(experiment, db) %>%
  get_residual_enrichments() %>%
  ggplot(aes(x=residual_beta, y=-log10(residual_p), color=result)) +
  geom_point() +
  scale_color_manual(values = c('depleted' = 'coral',
                                'enriched' = 'dodgerblue',
                                'not significant' = 'grey')) +
  facet_wrap(vars(db))
```

## Enrichment results

### Glossary

* `alpha` is the posterior probability of SuSiE including this gene set **in this component** which is different from PIP (probability of SuSiE including this gene set in ANY component)
* `beta` posterior mean/standard error of posterior mean for effect size. Standard errors are likely too small. 
* `oddsRatio, pHypergeometric, pFishersExact` construct a contingency table (gene list membersip) x (gene set membership), estimate the `oddsRatio` gives the odds of being in the gene list conditional on being in the gene set / odds of being in the gene list conditional on NOT being in the gene set. `pHypergeometric` and `pFishersExact` are pvalues from 1 and 2 sided test respectively. 


```{r}
#' @param tbl output of \alias{get_credible_set_summary} to be formatted
#' @param target_coverage the coverage of the credible sets to be reported
#' @param max_coverage report SNPs that are not in the target_coverage c.s. up this value
#' @param max_sets maximum number of gene sets to report for a single credible set
#'    this is useful for large credible sets
#' @return A styled kable table suitable for rendering in an HTML document
#' @export
report_susie_credible_sets = function(tbl,
                                      target_coverage=0.95,
                                      max_coverage=0.99,
                                      max_sets=10){
  require(kableExtra)
  tbl_filtered <-
    tbl %>%
    group_by(component) %>%
    arrange(component, desc(alpha)) %>%
    dplyr::filter(cumalpha <= max_coverage, alpha_rank <= max_sets) %>%
    dplyr::mutate(in_cs = cumalpha <= target_coverage) %>% 
    dplyr::ungroup() %>%
    dplyr::mutate(logOddsRatio = log10(oddsRatio))

  tbl_filtered %>%
    dplyr::select(component, geneSet, description, geneSetSize, overlap, logOddsRatio, beta, beta.se, alpha, pFishersExact) %>%
    dplyr::mutate_if(is.numeric, funs(as.character(signif(., 3)))) %>%
    pack_group %>%
    column_spec(c(5, 6), color=ifelse(tbl_filtered$beta > 0, 'green', 'red')) %>%
    kableExtra::kable_styling()
}
```

```{r experiment.level.results, results='asis'}
experiments <- unique(res$experiment)

do.experiment.volcano = function(this_experiment){
  res %>% 
    filter(experiment == this_experiment) %>%
    group_by(db) %>%
    get_ora_enrichments %>%
    ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
    geom_point() +
    geom_point(
      csdat %>% filter(experiment == this_experiment), 
      mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact), fill=component),
      color='black', pch=21, size=5) +
    scale_color_manual(values = c('depleted' = 'coral',
                                  'enriched' = 'dodgerblue',
                                  'not significant' = 'grey')) +
    facet_wrap(vars(db)) +
    labs(title = this_experiment)
}

do.residual.volcano = function(this_experiment){
  res %>% 
    filter(experiment == this_experiment) %>%
    group_by(db) %>%
    get_residual_enrichments() %>%
    ggplot(aes(x=residual_beta, y=-log10(residual_p), color=result)) +
    geom_point() +
    scale_color_manual(values = c('depleted' = 'coral',
                                  'enriched' = 'dodgerblue',
                                  'not significant' = 'grey')) +
    facet_wrap(vars(db)) +
    labs(title = paste0('Residual gene set associations: ', this_experiment))
}

for (i in 1:length(experiments)){
  this_experiment <- experiments[i]
  cat("\n") 
  cat("###", this_experiment, "\n") # Create second level headings with the names.
  volcano1 <- do.experiment.volcano(this_experiment) 
  volcano2 <- do.residual.volcano(this_experiment)
  print(volcano1)
  print(volcano2)
  cat("\n\n") 

  avail_dbs <- csdat %>% 
    filter(experiment == this_experiment) %>%
    select(db) %>% unique()
  for(this_db in avail_dbs){
    cat("####", this_db, "\n") # Create second level headings with the names.
    res2 %>%
      filter(
        active_cs,
        experiment == this_experiment,
        db == this_db) %>%
      distinct() %>%
      report_susie_credible_sets(max_coverage=0.95) %>% 
      htmltools::HTML() %>% 
      print()
    cat("\n")
  }}
```

### Enrichment table

```{r full.table}
color_sign = function(value) {
      if (value > 0) {
        color <- "#008000"
      } else if (value < 0) {
        color <- "#e00000"
      } else {
        color <- "#777"
      }
      list(color = color)
}

in_cs <-
  res2 %>% 
  filter(in_cs, active_cs) %>%
  select(experiment, db, thresh, geneSet, component, in_cs) %>%
  distinct()

dt <- res %>% 
  filter(overlap > 0) %>%
  mutate(
    logOddsRatio = log(oddsRatio),
    nlog10pFishersExact = -log10(pFishersExact)
  ) %>%
  left_join(in_cs) %>%
  mutate(in_cs = if_else(is.na(in_cs), FALSE, in_cs)) %>%
  select(in_cs, experiment, db, thresh, geneSet, beta, pip, overlap, geneSetSize, logOddsRatio, nlog10pFishersExact) %>%
  mutate(across(!where(is.numeric) , as.factor)) 

dt %>%
  select(-in_cs) %>%
  reactable::reactable(
    filterable=TRUE,
    minRows=10,
    columns = list(
      pip = reactable::colDef(format = reactable::colFormat(digits = 3)),
      logOddsRatio = reactable::colDef(style= function(value){color_sign(value)},
                                       format = reactable::colFormat(digits = 3)),
      beta = reactable::colDef(style= function(value){color_sign(value)},
                               format = reactable::colFormat(digits = 3)),
      nlog10pFishersExact = reactable::colDef(format = reactable::colFormat(digits = 3))
    ),
    rowStyle = function(index){
      if(dt$in_cs[index] == TRUE){
        list(background = "#e5f5e0")
      }
    },
    defaultSorted = list(nlog10pFishersExact='desc')
  )
```
