---
title: "Chondrocyte CTS and IL1B DE"
author: "karltayeb"
date: "2022-04-11"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

## Introduction

Anthony differentiated IPSCS -> MSC -> Chondrocytes and treated with physical stress and IL1B.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '#>',
  message=FALSE, warning=FALSE
)
```


## Load data

```{r}
source('code/load_data.R')
library(gseasusie)

# save output to a folder... helps for running many factors
cache_rds = purrr::partial(xfun::cache_rds, dir='cache/chondrocyte_oa/')

driver = function(gs, experiment){
  dat <- gseasusie::prep_binary_data(gs, data[[experiment]], thresh = 0.05)
  ora <- gseasusie::fit_ora(dat$X, dat$y)
  # add description
  ora <- ora %>%
    dplyr::left_join(gs$geneSetDes)
  fit <- gseasusie::fit_logistic_susie_veb_boost(dat$X, dat$y)
  return(list(fit=fit, ora=ora))
}

driver_cached = function(prefix, gs, experiment, ...){
  file = paste0(prefix, '_', gs$db, '_', experiment, '.rds')
  print(file)
  cache_rds(driver(gs, experiment), file=file, ...)
}
```

```{r fits}
data  <- load_chondrocyte_data2()
go <- gseasusie::load_all_go()
res <- purrr::map(names(data), ~driver_cached('hnscc', go, .x, rerun=F))
names(res) <- names(data)
```

```{r, eval=F}
pack_group = function(tbl){
    components <- tbl$component
    unique.components <- unique(components)
    start <- match(unique.components, components)
    end <- c(tail(start, -1) - 1, length(components))
    res <- tbl %>% dplyr::select(-c(component)) %>% kableExtra::kbl()
    for(i in 1:length(unique.components)){
      res <- kableExtra::pack_rows(res, unique.components[i], start[i], end[i])
    }
    return(res)
}

#' Report credible set based summary of SuSiE
#' @export
static_table = function(fit, ora){
  # get credible sets
  res <- gseasusie:::get_gene_set_summary(fit) %>%
    dplyr::left_join(ora)
  csdat <- gseasusie:::get_credible_set_summary(fit) %>%
    dplyr::left_join(ora) %>%
    dplyr::filter(in_cs, active_cs) %>%
    dplyr::select(geneSet, description, component, in_cs, alpha, conditional_beta) %>%
    distinct()
  
  # manipulate table
  columns <- c(
    'geneSet', 'description', 'geneSetSize', 'overlap',
    'log2OR', 'effect', 'alpha', 'pip', 'nlog10pFishersExact', 'fisherRank',
    'component'
  )
  color_columns <- which(columns %in% c('log2OR', 'effect'))
  dt <- res %>% 
    dplyr::filter(overlap > 0) %>%
    dplyr::mutate(
      log2OR = log2(oddsRatio),
      nlog10pFishersExact = -log10(pFishersExact)
    ) %>%
    dplyr::left_join(csdat) %>%
    dplyr::arrange(dplyr::desc(nlog10pFishersExact)) %>%
    dplyr::mutate(
      fisherRank = dplyr::row_number(),
      in_cs = dplyr::if_else(is.na(in_cs), FALSE, in_cs),
      effect = conditional_beta * log2(exp(1))
    ) %>%
    dplyr::filter(in_cs) %>%
    dplyr::select(columns) %>%
    dplyr::mutate(dplyr::across(!where(is.numeric) , as.factor)) %>%
    mutate(
      component = reorder(factor(component), fisherRank, FUN=min) 
      # sorts components
    ) %>%
    dplyr::arrange(component) %>%
    dplyr::mutate_if(is.numeric, funs(as.character(signif(., 3))))

  # display table
  dt %>%
    pack_group %>%
    kableExtra::column_spec(
      color_columns, color=ifelse(dt$effect > 0, 'green', 'red')) %>%
    kableExtra::kable_styling()
}

static_table(fit, ora)
```


```{r hnscc.prep, message=FALSE, warning=FALSE}
results <- res
experiments <- names(results)

safe_static_table <- purrr::safely(
  gseasusie::static_table,
  otherwise = 'nothing to report')

for(this_experiment in experiments){
  sub_res <- results[[this_experiment]]
  volcano <- gseasusie::enrichment_volcano(fit=sub_res$fit, ora=sub_res$ora)
  results[[this_experiment]]$volcano <- volcano
  
  table <- safe_static_table(sub_res$fit, sub_res$ora)$result
  results[[this_experiment]]$table <- table
}
```


```{r results, results='asis'}
cat('\n')
cat('##', 'Results')

for(this_experiment in experiments){
  cat('\n')
  cat('###', this_experiment)

  #sub_res <- results[[this_experiment]]
  #volcano <- gseasusie::enrichment_volcano(fit=sub_res$fit, ora=sub_res$ora)
  print(results[[this_experiment]]$volcano)
  
  cat("\n\n")
  #table <- safe_static_table(sub_res$fit, sub_res$ora)$result
  print(results[[this_experiment]]$table)
}
```

```{r}
knitr::knit_exit()
```


```{r load.results}
library(gseasusie)
library(ggplot2)
library(tidyverse)
library(targets)

source('code/target_run_enrichment.R')

#tar_load(genesets)
#tar_load(chondroctye_oa_meta)
#tar_load(chondroctye_oa_data)
tar_load(chondroctye_oa_enrichment_fit)

results <- tar_read(chondroctye_oa_enrichment_summary)
```

## Enrichment results

```{r}
pack_group = function(tbl){
    components <- tbl$component
    unique.components <- unique(components)
    start <- match(unique.components, components)
    end <- c(tail(start, -1) - 1, length(components))
    res <- tbl %>% dplyr::select(-c(component)) %>% kableExtra::kbl()
    for(i in 1:length(unique.components)){
      res <- kableExtra::pack_rows(res, unique.components[i], start[i], end[i])
    }
    return(res)
}


static_table2 = function(res){
  require(kableExtra)
  tbl_filtered <-
    res %>%
    arrange(pFishersExact) %>%
    mutate(fisherRank = row_number()) %>%
    filter(in_cs, active_cs) %>%
    group_by(component) %>%
    arrange(component, desc(alpha)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(logOddsRatio = log(oddsRatio))

  tbl_filtered %>%
    dplyr::select(
      component, geneSet, description, geneSetSize, overlap,
      logOddsRatio, conditional_beta, conditional_beta_se,
      alpha, pip, pFishersExact, fisherRank) %>%
    dplyr::mutate_if(is.numeric, funs(as.character(signif(., 3)))) %>%
    pack_group %>%
    column_spec(c(5, 6), color=ifelse(tbl_filtered$beta > 0, 'green', 'red')) %>%
    kableExtra::kable_styling()
}
```

```{r experiment.level.results, results='asis'}
results <- results %>% arrange(experiment, ptop, db)
experiments <- results$experiment %>% unique()

for(this_experiment in experiments){
  cat('\n')
  cat('##', this_experiment)

  sub_res <- results %>% filter(experiment == this_experiment)

  N <- dim(sub_res)[1]
  for (i in 1:N){
    this_db <- sub_res$db[[i]]
    this_ptop <- sub_res$ptop[[i]]
    res <- sub_res$enrichment_summary[[i]]

    cat("\n") 
    cat("### prop genes = ", this_ptop, ', db = ', this_db, "\n") # Create second level headings with the names.
    volcano <- res %>% gseasusie::enrichment_volcano2()
    hist <- res %>% filter(overlap > 0) %>% gseasusie::residual_enrichment_histogram2() 
    print(volcano)
    print(hist)
    cat("\n\n")
    print(safely(static_table2, NULL, quiet = TRUE)(res))
  }
}
```

### Glossary

* `beta` $\mathbb E_q[\beta_l | \gamma_l = j]$  the posterior mean effect size, conditional on this gene set ($j$) being selected by component $l$
* `alpha` $\mathbb E_q(\gamma_l = j)$ is the posterior probability of SuSiE selecting this gene set ($j$). Not this is distinct from the PIP $\text{PIP}_j = 1 - \prod_l (1 - \alpha_{lj})$
* `beta` posterior mean/standard error of posterior mean for effect size. Standard errors are likely too small. 
* `oddsRatio, pHypergeometric, pFishersExact` construct a contingency table (gene list membersip) x (gene set membership), estimate the `oddsRatio` gives the odds of being in the gene list conditional on being in the gene set / odds of being in the gene list conditional on NOT being in the gene set. `pHypergeometric` and `pFishersExact` are pvalues from 1 and 2 sided test respectively. 
