---
title: "Getting started"
author: "karltayeb"
date: "2022-05-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = F, message = F
)
```


## Introduction

Here we demonstrate the basics of how to use the `gseasusie` package.

```{r, eval=FALSE}
devtools::install_github('karltayeb/gseasusie')
```


```{r, message=FALSE, results='hide'}
library(gseasusie)
library(tidyverse)
```

## Loading gene sets
`gseasusie` ships with lots of gene set databases!
There are other packages that curate geneset databases `WebGestaltR`, `msigdbr`, and `stephenslab/pathways`
All I did here was write a few functions to load them into a standard format.

Each gene set list contains three elements: 
 * `X`: a gene x gene set indicator matrix, genes are identified by their ENTREZID
 * `geneSet`: a two column `data.frame` (`geneSet` and `gene`) mapping genes to gene sets 
 * `geneSetDes`: a written description of what the gene set does
 
`gseasusie::load_gene_sets` lets you load multiple gene sets. Note: the first time you try to load gene sets it might take a while because we need to build the indicator matrices (and the implimentation is far from optimized). After you load them once it will cache to a folder `./cache/resources/` where the path is relative to your working directory/project directory.

```{r load.gene.sets}
genesets <- gseasusie::load_gene_sets(c('c2', 'all_msigbd'))
genesets$c2$geneSet %>% head()
genesets$c2$geneSetDes %>% head()
dim(genesets$c2$X)
```

## How to format data

### Managing multiple experiment

If you have multiple experiments that you want to run enrichment for, I've found it useful to format as follows.
`data` is a list of data frames, one for each experiment. The dataframes can have whatever information you'd like, but they must have the following columns: 
* `ENTREZID`: the gene sets above use `ENTREZID`, so map your gene names to this!
* `threshold.on`: this might be a p-value, adjusted pvalue, effect size, etc
* and `beta`:: right now this is only used for sign information, so if you don't care specifically about up or down regulated genes just make a dummy column with 1s. This will be more important once we support enrichment on z-scores and effects sizes without thresholding.

```{r load.data}
source('code/load_data.R')
data <- load_sc_pbmc_deseq2()
data$`CD19+ B` %>% head()
```


### Binarizing the data

At the end of the day, you're going to need a binary gene x gene set matrix `X` and a binary gene list `y`. However you want to get those is fine.

One benefit of organizing your data as in the previous section, `gseasusie` ships with some helpful functions to format/prepare the data. `gseasusie::prep_binary_data` will binarize the data by thresholding on `threshold.on`.

```{r binarize.data}
db <- 'c2'  # name of gene set database to use in `genesets`
experiment = 'CD19+ B'  # name of experiment to use in `data`
thresh = 1e-4  # threshold for binarizing the data
bin.data <- gseasusie::prep_binary_data(genesets[[db]], data[[experiment]], thresh)

X <- bin.data$X
y <- bin.data$y
```


## Fitting different enrichment models/methods

Now we can fit our enrichment models.
NOTE: The marginal regressions are implemented in python (`basilisk` will spin up a conda environment with necessary dependencies, which will take some time the first time you run it, but will run quickly after!)

```{r fit.models}
# fit logistic susie
logistic.fit <- gseasusie::fit_logistic_susie_veb_boost(X, y, L=20)

# fit linear susie
# (all of the functions that work with logistic susie fits should work with regular susie fits)
linear.fit <- susieR::susie(X, y)

# compute odds ratios, and pvalues under hypergeometric (one-sided) and fishers exact (two-sided) tests
ora <- gseasusie::fit_ora(X, y)

# like ORA but performed implimented as a univariate logistic regression
marginal_regression <- gseasusie::fit_marginal_regression_jax(X, y)

# redo the logistic regression ORA, conditional on enrichments found by logistic susie
# in practice, we introduce logistic susie predictions as an offset in the new model
residual_regression <- gseasusie::fit_residual_regression_jax(X, y, logistic.fit)

res = list(
  fit=logistic.fit,
  ora=ora,
  marginal_reg = marginal_regression,
  residual_reg = residual_regression
)
```


## Visualizations

We can visualize our results with a volcano plot (more plots coming)
The large circles highlight gene sets in a 95% credible set. The color indicates which SuSiE component)

```{r make.volcano}
gseasusie::enrichment_volcano(logistic.fit, ora)
```

If we account for the predictions made by (logitic) SuSiE, we can see that a lot (but not all of) the enrichment signal has been accounted for.

```{r make.residual.hist}
gseasusie::residual_enrichment_histogram(marginal_regression, residual_regression)
```

## Interactive tables

We can produce an interactive table to explore the enrichment results:

```{r make.itable}
gseasusie::interactive_table(logistic.fit, ora)
```


