---
title: "gseabenchmark_tcga"
author: "karltayeb"
date: "2022-03-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Our goals here are to run Logistic SuSiE on differential expression results from TCGA.
We want to assess:
1. If the resulting enrichment results look good/interpretable across multiple/concatenated gene sets
2. Assess sensitivity to a range of p-value thresholds
3. Evaluate the potential of the summary stat latent model

```{r}
library(GSEABenchmarkeR)
library(EnrichmentBrowser)
library(tidyverse)
source('code/load_gene_sets.R')
```


#### Load Gene Sets

`loadGeneSetX` uniformly formats gene sets and generates the $X$ matrix
We can source any gene set from `WebGestaltR::listGeneSet()`

```{r}
gs_list <- WebGestaltR::listGeneSet()

gobp <- loadGeneSetX('geneontology_Biological_Process', min.size=50)  # just huge number of gene sets
gobp_nr <- loadGeneSetX('geneontology_Biological_Process_noRedundant', min.size=1)
gomf <- loadGeneSetX('geneontology_Molecular_Function', min.size=1)
kegg <- loadGeneSetX('pathway_KEGG', min.size=1)
reactome <- loadGeneSetX('pathway_Reactome', min.size=1)
wikipathway_cancer <- loadGeneSetX('pathway_Wikipathway_cancer', min.size=1)
wikipathway <- loadGeneSetX('pathway_Wikipathway', min.size=1)

gobp <- load.webGestalt.geneSet('geneontology_Biological_Process')
X.gobp <- gobp %>% convertGeneSet(min.size=50) %>% geneSet2X

gobp <- load.webGestalt.geneSet('geneontology_Biological_Process')
X.gobp <- gobp %>% convertGeneSet(min.size=50) %>% geneSet2X
```


#### Load TCGA Data
```{r}
# get z scores
tcga <- loadEData("tcga", nr.datasets=2)
tcga <- runDE(tcga, de.method='limma', padj.method="none")
```

#### Get P
```{r}
df <- rowData(tcga$BRCA)
rowData(tcga$BRCA)$z <- -1 * qnorm(rowData(tcga$BRCA)$PVAL) * sign(rowData(tcga$BRCA)$FC)
z <- df$z %>% `names<-`(rownames(df))
pvals <- df$PVALUE %>% `names<-`(rownames(df))
```


```{r}
# get z scores
tcga <- loadEData("tcga", nr.datasets=2)
tcga <- runDE(tcga, de.method='limma', padj.method="none")
rowData(tcga$BRCA)$z <- -1 * qnorm(rowData(tcga$BRCA)$PVAL) * sign(rowData(tcga$BRCA)$FC)
df <- rowData(tcga$BRCA)
z <- df$z %>% `names<-`(rownames(df))

# load gene sets
gs2X <- function(gs){
  unique.genes <- unique(unlist(gs))
  X <- purrr::map(gs, ~ Matrix::Matrix(unique.genes %in% .x %>% as.integer, sparse = T)) %>%
    Reduce(cbind2, .) %>%
    `rownames<-`(unique.genes) %>%
    `colnames<-`(names(gs))
  return(X)
}
kegg.gs <- getGenesets(org="hsa", db="kegg")
kegg.gs.info <- getGenesets(org="hsa", db="kegg", return.type = 'GeneSetCollection')

X.kegg <- gs2X(kegg.gs)
```

```{r}
source('code/utils.R')
source('code/logistic_susie_vb.R')
source('code/logistic_susie_veb_boost.R')
source('code/latent_logistic_susie.R')
library(susieR)

u <- process_input(X.kegg, z)
X <- u$X
z <- u$y
y3 <- as.integer(abs(z) > 3) %>% `names<-`(names(z))
y10 <- as.integer(abs(z) > 10) %>% `names<-`(names(z))

y.susie.fit <- susie(X, y10)
z.susie.fit <- susie(X, z)
veb.fit <- logistic.susie.veb.boost(X, y10, L=10)
vb.fit <- logistic.susie(X, y10, L=10)

source('code/latent_logistic_susie.R')
f0 <- function(x){dnorm(x, mean=0, sd=1, log=T)}
f1 <- function(x){dnorm(x, mean=0, sd=10, log=T)}
z.vb.fit <- latent.logistic.susie(X, z, f0, f1, maxit=10, verbose=1)

par(mfrow=c(2,3))
susie_plot(y.susie.fit, "PIP", main='susie z score')
susie_plot(z.susie.fit, "PIP", main='susie binary')
plot.new()

susie_plot(veb.fit, "PIP", main='veb.boost')
susie_plot(vb.fit, "PIP", main='logistic SuSiE')
susie_plot(z.vb.fit, "PIP", main = 'latent logistic SuSiE')
```

```{r}
source('code/logistic_susie_vb.R')
vb.fit <- logistic.susie(X, y10, L=10, init.intercept=0, intercept = T, maxit=10)
vb.fit$intercept

vb.fit <- logistic.susie(X, y10, L=50, init.intercept=-5, intercept=F, maxit=100)
susie_plot(vb.fit, 'PIP')
```

## Vary z-score/ p-value threshold

```{r}
thresh <- c(2, 4, 6, 8, 10, 12)
names(thresh) <- paste0('z', thresh)

veb.thresh <- lapply(thresh, function(t){logistic.susie.veb.boost(X, as.integer(abs(z) > t), L=10)})
vb.thresh <- lapply(thresh, function(t){logistic.susie(X, as.integer(abs(z) > t), L=10, maxit=100)})

par(mfrow=c(2,3))
for(t in names(thresh)){
  susie_plot(vb.thresh[[t]], "PIP", main=t)
}
```


### Vary f0, f1

```{r}
hist(-log10(df$PVAL))
hist(df$PVAL)
hist()

```


```{r}
z.vb.fit$intercept

mean(y3)
```



### getting z score model to work

```{r}

```

