---
title: "Human Chimp EB"
author: "karltayeb"
date: "2022-04-11"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
params:
  cache_dir: 'cache/human_chimp_eb/'
  genesets: c('gobp', 'gomf', 'kegg')
  thresh: 1e-2
  rerun: TRUE  # set to true to ignore cache
---

```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = '/project2/gilad/ktayeb/logistic-susie-gsea/'
)
knitr::opts_chunk$set(
  comment = '',
  message=FALSE, warning=FALSE,
  root.dir = '/project2/gilad/ktayeb/logistic-susie-gsea/'
)
```

## Introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '#>',
  message=FALSE, warning=FALSE
)
```

```{r load.libraries, include=FALSE}
library(tidyverse)
library(htmltools)
library(gseareg)
source('code/utils.R')


params = list(
  cache_dir = 'cache/human_chimp_eb/',
  genesets=('kegg'), thresh=1e-2, rerun=FALSE
)
params.cache_dir <- params$cache_dir
params.rerun <- eval(parse(text=params$rerun))
cache_rds <- partial(xfun::cache_rds, dir=params.cache_dir, rerun=params.rerun)
```

```{r load.data, include=FALSE}
source('code/load_data.R')
genesets <- gseareg::load_gene_sets()
data <- load_human_chimp_eb()
```

```{r}
pvals <- map(data, ~pluck(.x, 'threshold.on'))
hist(map_dbl(pvals, ~mean(.x[!is.na(.x)] < 1e-2)), main='proportion of "interesting" across cell-types')
```

```{r fit, include=FALSE}
params.genesets <- eval(parse(text=params$genesets))
params.thresh <- eval(parse(text=params$thresh))

# fit logistic susie
do_logistic_susie_cached = function(data,
                                    db,
                                    thresh,
                                    prefix=''){
  res <- cache_rds({
    purrr::map_dfr(
      names(data),
      ~do_logistic_susie(.x, db, thresh, genesets, data))
  },
  file=paste0(prefix, 'logistic_susie_', db, '_', thresh),
  hash = list(data, db, thresh, prefix))
}
fits <- map_dfr(params.genesets, ~do_logistic_susie_cached(data, .x, params.thresh))

do_linear_susie_cached = function(data,
                                    db,
                                    thresh,
                                    prefix=''){
  res <- cache_rds({
    purrr::map_dfr(
      names(data),
      ~gseareg::do_linear_susie(.x, db, thresh, genesets, data))
  },
  file=paste0(prefix, 'logistic_susie_', db, '_', thresh),
  hash = list(data, db, thresh, prefix))
}
fits <- map_dfr(params.genesets, ~do_linear_susie_cached(data, .x, params.thresh))

# tccm
do_tccm_susie_cached = function(data,
                                    db,
                                    thresh,
                                    prefix=''){
  res <- cache_rds({
    purrr::map_dfr(
      names(data),
      ~gseareg::do_tccm_point_normal_susie(.x, db, thresh, genesets, data))
  },
  file=paste0(prefix, 'logistic_susie_', db, '_', thresh),
  hash = list(data, db, thresh, prefix))
}
fits <- map_dfr(params.genesets, ~do_linear_susie_cached(data, .x, params.thresh))

# fit ora
do_ora_cached = function(data, db, thresh, prefix='', ...){
  res <- cache_rds({
    purrr::map_dfr(
      names(data),
      ~do_ora(.x, db, thresh, genesets, data))
  },
  file=paste0(prefix, 'ora_', db, '_', thresh),
  hash = list(data, db, thresh, prefix))
}
ora <- map_dfr(params.genesets, ~do_ora_cached(data, .x, params.thresh))
```

## Overview
### Threshold sensitivity

```{r threshold.sensitivity}
ttt <- map_dbl(1:10, ~10**-.x)

.prop.ones = function(experiment){
  map_dbl(ttt, ~ prep_binary_data(
  genesets[['gobp']], data[[experiment]], thresh=.x)$y %>% mean())
}

prop.ones <- cache_rds({map_dfc(names(data), ~.prop.ones(.x))}, file='threshold_sensitivity')
colnames(prop.ones) <- names(data)
prop.ones <- prop.ones %>% mutate(thresh = ttt)

prop.ones %>%
  pivot_longer(one_of(names(data))) %>%
  group_by(name) %>%
  mutate(value = value) %>%
  ggplot(aes(x=factor(-log10(thresh)), y=value)) +
  geom_boxplot() +
  labs(
    y = 'Proportions of genes in gene list',
    title = 'Sensitivity to thresholding'
  )
```

```{r process.results, echo=FALSE, warning=FALSE}
# gene set summary
res <- fits %>%
  left_join(ora) %>%
  mutate(
    gs_summary = map(fit, get_gene_set_summary),
    res = map2(gs_summary, ora, ~ left_join(.x, .y)),
  ) %>%
  select(-c(fit, ora, gs_summary)) %>%
  unnest(res)

# credible set summary
res2 <- fits %>%
  left_join(ora) %>%
  mutate(res = map(fit, get_credible_set_summary)) %>%
  mutate(res = map2(res, ora, ~ left_join(.x, .y))) %>%
  select(-c(fit, ora)) %>%
  unnest(res)

# Setup HTML tables
split_tibble = function(tibble, col = 'col'){
  tibble %>% split(., .[, col])
}
descriptions <- map_dfr(genesets, ~pluck(.x, 'geneSet', 'geneSetDes'))
tbl <- res2 %>% 
  filter(active_cs) %>%
  left_join(descriptions)
tbl_split <- split_tibble(tbl, 'experiment')
html_tables <- map(tbl_split, ~split_tibble(.x, 'db'))
```


### Big volcano plot

Colors represent enrichment/depletion detected by Fishers exact test (Benjamini Hochberg corrected p-values < $0.05$). Gene sets that belong to a SuSiE credible set are circled.

```{r big.volcano.plot}
get_ora_enrichments = function(tbl){
   tbl %>% mutate(
    padj = p.adjust(pFishersExact),
    result = case_when(
      padj < 0.05 & oddsRatio < 1 ~ 'depleted',
      padj < 0.05 & oddsRatio > 1 ~ 'enriched',
      TRUE ~ 'not significant'
    )
  )
}

# plot all enrichments, highlight gene sets in credible set
csdat <- res2 %>% 
  filter(in_cs, active_cs)

res %>% 
  group_by(experiment, db) %>%
  get_ora_enrichments %>%
  ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
  geom_point() +
  geom_point(
    csdat, 
    mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact)),
    color='black', pch=21, size=5) +
  scale_color_manual(values = c('depleted' = 'coral',
                                'enriched' = 'dodgerblue',
                                'not significant' = 'grey')) +
  facet_wrap(vars(db))
```

## Enrichment results

### Glossary

* `alpha` is the posterior probability of SuSiE including this gene set **in this component** which is different from PIP (probability of SuSiE including this gene set in ANY component)
* `beta` posterior mean/standard error of posterior mean for effect size. Standard errors are likely too small. 
* `oddsRatio, pHypergeometric, pFishersExact` construct a contingency table (gene list membersip) x (gene set membership), estimate the `oddsRatio` gives the odds of being in the gene list conditional on being in the gene set / odds of being in the gene list conditional on NOT being in the gene set. `pHypergeometric` and `pFishersExact` are pvalues from 1 and 2 sided test respectively. 

```{r experiment.level.results, results='asis'}
experiments <- unique(res$experiment)

do.experiment.volcano = function(this_experiment){
  res %>% 
    filter(experiment == this_experiment) %>%
    group_by(db) %>%
    get_ora_enrichments %>%
    ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
    geom_point() +
    geom_point(
      csdat %>% filter(experiment == this_experiment), 
      mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact)),
      color='black', pch=21, size=5) +
    scale_color_manual(values = c('depleted' = 'coral',
                                  'enriched' = 'dodgerblue',
                                  'not significant' = 'grey')) +
    facet_wrap(vars(db)) +
    labs(title = this_experiment)
}

for (i in 1:length(experiments)){
  this_experiment <- experiments[i]
  cat("\n") 
  cat("###", this_experiment, "\n") # Create second level headings with the names.
  do.experiment.volcano(this_experiment) %>% print()
  cat("\n\n") 

  for(db in names(html_tables[[this_experiment]])){
    cat("####", db, "\n") # Create second level headings with the names.
    to_print <- html_tables[[this_experiment]][[db]] %>% distinct()
    to_print %>% report_susie_credible_sets() %>% htmltools::HTML() %>% print()
    cat("\n")
  }}
```

### Enrichment table

```{r full.table}
color_sign = function(value) {
      if (value > 0) {
        color <- "#008000"
      } else if (value < 0) {
        color <- "#e00000"
      } else {
        color <- "#777"
      }
      list(color = color)
}

in_cs <-
  res2 %>% 
  filter(in_cs, active_cs) %>%
  select(experiment, db, thresh, geneSet, component, in_cs) %>%
  distinct()

dt <- res %>% 
  filter(overlap > 0) %>%
  mutate(
    logOddsRatio = log(oddsRatio),
    nlog10pFishersExact = -log10(pFishersExact)
  ) %>%
  left_join(in_cs) %>%
  mutate(in_cs = if_else(is.na(in_cs), FALSE, in_cs)) %>%
  select(in_cs, experiment, db, thresh, geneSet, beta, pip, overlap, geneSetSize, logOddsRatio, nlog10pFishersExact) %>%
  mutate(across(!where(is.numeric) , as.factor)) 

dt %>%
  select(-in_cs) %>%
  reactable::reactable(
    filterable=TRUE,
    minRows=10,
    columns = list(
      pip = reactable::colDef(format = reactable::colFormat(digits = 3)),
      logOddsRatio = reactable::colDef(style= function(value){color_sign(value)},
                                       format = reactable::colFormat(digits = 3)),
      beta = reactable::colDef(style= function(value){color_sign(value)},
                               format = reactable::colFormat(digits = 3)),
      nlog10pFishersExact = reactable::colDef(format = reactable::colFormat(digits = 3))
    ),
    rowStyle = function(index){
      if(dt$in_cs[index] == TRUE){
        list(background = "#e5f5e0")
      }
    },
    defaultSorted = list(nlog10pFishersExact='desc')
  )
```
