---
title: "Comparing linear and logistic SuSiE for GSEA"
author: "karltayeb"
date: "2022-01-06"
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Where Linear SuSiE Fails

Linear SuSiE successfully select the true active gene set. However, when the active gene set is only partially observed in the list (in this case, genes in the active gene set are observed with probability $\frac{1}{2}$), and there are not many background genes included in the list, SuSiE will tend to include many irrelevant gene sets. I claim the linear model finds it advantageous to include extra gene sets to help offset the positive predictions on active genes that are unobserved in the gene list. In contrast logistic SuSiE correctly selects the active gene set without the inclusion of false positives.

### Questions that come to mind

- If our claim is true, then we should expect the false positive gene sets to have negative coefficients. This seems true in the case of `background.logit=-6`, but not for `background.logit=-10`.
- Broadly, when is linear regression a good approximation to logistic regression. Does intuition in the univariate setting hold in the multivariate setting? In a multivariate setting with correlated covariates?
- Fully observed gene sets are not susceptible to this pitfall in linear SuSiE. At what rate do active genes need to be included in the list to avoid this pitfall? It is probably a function of the gene set size, the background inclusion rate, and the degree of overlap between the active gene set and other gene sets.
- We observe this result using GO-BP no-redundant gene set, thus there is substantially less overlap among gene sets considered by the model. How do linear and logistic SuSiE compare when gene sets are more/less correlated?

```{r, linear.v.logistic.simulation, class.source = 'fold-hide'}
library(tidyverse)
source('code/load_gene_sets.R')
source('code/simulate_gene_lists.R')
source('code/fit_susie.R')
source('code/fit_logistic_susie.R')

X <- load.webGestalt.X()

set.seed(0)
sim10 <- simulate.constant.sim(
      X, L=1,
      background.logit=-10,
      active.logit=0)
susie.10 <- fit.susie(X, sim10$Y[, 1], L=10)
logistic.susie.10 <- fit.logistic.susie.veb.boost(X, sim10$Y[, 1], k=10)


sim6 <- simulate.constant.sim(
      X, L=1,
      background.logit=-6,
      active.logit=0)
susie.6 <- fit.susie(X, sim6$Y[, 1], L=10)
logistic.susie.6 <- fit.logistic.susie.veb.boost(X, sim6$Y[, 1], k=10)


sim2 <- simulate.constant.sim(
      X, L=1,
      background.logit=-2,
      active.logit=0)
susie.2 <- fit.susie(X, sim2$Y[, 1], L=10)
logistic.susie.2 <- fit.logistic.susie.veb.boost(X, sim2$Y[, 1], k=10)
```

```{r, linear.v.logistic.pips, out.width='75%', fig.align='center', class.source = 'fold-hide'}
plot.pip <- function(sim, model, ...){
  plot(model$pip[[1]], ...)
  points(which(sim$active !=0), model$pip[[1]][sim$active != 0], col='red')
}

par(mfrow=c(2, 3), mai=c(.5,.5,.3,.3), oma=c(1.5,1.5,1.5,.5))

plot.pip(sim10, susie.10, main='SuSiE PIP background.logit=-10')
plot.pip(sim6, susie.6, main='SuSiE PIP background.logit=-6')
plot.pip(sim2, susie.2, main='SuSiE PIP background.logit=-2')

plot.pip(sim10, logistic.susie.10, main='Logistic SuSiE PIP background.logit=-10')
plot.pip(sim6,  logistic.susie.6, main='Logistic SuSiE PIP background.logit=-6')
plot.pip(sim2,  logistic.susie.2, main='Logistic SuSiE PIP background.logit=-2')
```


```{r, linear.v.logistic.coef, out.width='75%', fig.align='center'}
plot.coef <- function(sim, model, ...){
  plot(model$coef[[1]], ...)
  points(which(sim$active !=0), model$coef[[1]][sim$active != 0], col='red')
}

par(mfrow=c(2, 3), mai=c(.5,.5,.3,.3), oma=c(1.5,1.5,1.5,.5))

plot.coef(sim10, susie.10, main='SuSiE PIP background.logit=-10')
plot.coef(sim6, susie.6, main='SuSiE PIP background.logit=-6')
plot.coef(sim2, susie.2, main='SuSiE PIP background.logit=-2')

plot.coef(sim10, logistic.susie.10, main='Logistic SuSiE PIP background.logit=-10')
plot.coef(sim6,  logistic.susie.6, main='Logistic SuSiE PIP background.logit=-6')
plot.coef(sim2,  logistic.susie.2, main='Logistic SuSiE PIP background.logit=-2')
```



We can contrast the behavior with the case where the active gene set is fully observed:


```{r, linear.v.logistic.simulation2}
set.seed(1)
sim10 <- simulate.constant.sim(
      X, L=1,
      background.logit=-10,
      active.logit=10)
susie.10 <- fit.susie(X, sim10$Y[, 1], L=10)
logistic.susie.10 <- fit.logistic.susie.veb.boost(X, sim10$Y[, 1], k=10)


sim6 <- simulate.constant.sim(
      X, L=1,
      background.logit=-6,
      active.logit=10)
susie.6 <- fit.susie(X, sim6$Y[, 1], L=10)
logistic.susie.6 <- fit.logistic.susie.veb.boost(X, sim6$Y[, 1], k=10)


sim2 <- simulate.constant.sim(
      X, L=1,
      background.logit=-2,
      active.logit=10)
susie.2 <- fit.susie(X, sim2$Y[, 1], L=10)
logistic.susie.2 <- fit.logistic.susie.veb.boost(X, sim2$Y[, 1], k=10)
```


```{r, linear.v.logistic.pips2, out.width='75%', fig.align='center'}
plot.pip <- function(sim, model, ...){
  plot(model$pip[[1]], ...)
  points(which(sim$active !=0), model$pip[[1]][sim$active != 0], col='red')
}

par(mfrow=c(2, 3), mai=c(.5,.5,.3,.3), oma=c(1.5,1.5,1.5,.5))

plot.pip(sim10, susie.10, main='SuSiE PIP background.logit=-10')
plot.pip(sim6, susie.6, main='SuSiE PIP background.logit=-6')
plot.pip(sim2, susie.2, main='SuSiE PIP background.logit=-2')

plot.pip(sim10, logistic.susie.10, main='Logistic SuSiE PIP background.logit=-10')
plot.pip(sim6,  logistic.susie.6, main='Logistic SuSiE PIP background.logit=-6')
plot.pip(sim2,  logistic.susie.2, main='Logistic SuSiE PIP background.logit=-2')
```


```{r, linear.v.logistic.coef2, out.width='75%', fig.align='center'}
plot.coef <- function(sim, model, ...){
  plot(model$coef[[1]], ...)
  points(which(sim$active !=0), model$coef[[1]][sim$active != 0], col='red')
}

par(mfrow=c(2, 3), mai=c(.5,.5,.3,.3), oma=c(1.5,1.5,1.5,.5))

plot.coef(sim10, susie.10, main='SuSiE PIP background.logit=-10')
plot.coef(sim6, susie.6, main='SuSiEl PIP background.logit=-6')
plot.coef(sim2, susie.2, main='SuSiE PIP background.logit=-2')

plot.coef(sim10, logistic.susie.10, main='Logistic SuSiE PIP background.logit=-10')
plot.coef(sim6,  logistic.susie.6, main='Logistic SuSiE PIP background.logit=-6')
plot.coef(sim2,  logistic.susie.2, main='Logistic SuSiE PIP background.logit=-2')
```
