---
title: "redundancy_is_a_problem"
author: "karltayeb"
date: "2022-05-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  cache_dir: 'cache/redundancy/'
  rerun: FALSE  # set to true to ignore cache
---

## Introduction

This notebook is intended to help communicate the issue of redundancy in GSEA (and how multiple regression can help!)

We take ~2500 GO-BP terms and simulate a single active gene set (at different gene set sizes) The degree of redundant enrichments is expectedly a function of gene set size and overlap.

Next we generate a gene list with 3 enriched gene sets and show that the redundant marginal enrichment from the largest gene set "bury" the other interesting enticements.

Finally, we demonstrate the utility of joint gene set analysis. A simple forward selection procedure successfully recovers the "active" gene sets.

### Example

```{r}
library(tidyverse)
library(purrr)
library(gseasusie)

params.cache_dir <- params$cache_dir
params.rerun <- eval(parse(text=params$rerun))
cache_rds <- partial(xfun::cache_rds, dir=params.cache_dir, rerun=params.rerun)
```

```{r}
set.seed(4)
active_gene_sets <- c("GO:0050994", "GO:0044774", "GO:0090150")

genesets <- gseasusie::load_gene_sets()
X <- genesets$gobp$X
ngenes <- sort(Matrix::colSums(X))
overlap <- Matrix::t(X) %*% X
```

```{r}
sim_from_gene_set = function(x, p.in=0.5, p.out=0.01){
  1 - rbinom(length(x), size=1, prob=ifelse(x == 1, p.in, p.out))
}
y <- map(
  active_gene_sets,
  ~tibble(ENTREZID=rownames(X), beta=1, threshold.on=sim_from_gene_set(X[,.x]))
)
names(y) <- active_gene_sets

ora_res <- cache_rds({
  map_dfr(active_gene_sets, ~do_ora(
  .x, 'gobp', thresh = 0.5, genesets = genesets, data = y))},
  file='l1.sim')
```

```{r}
null.y <- map(active_gene_sets[1:1],
              ~tibble(ENTREZID=rownames(X), beta=1, threshold.on=sim_from_gene_set(X[,.x], p.in=0.01)))
names(null.y) <- 'null'
ora_null_res <- do_ora(
  'null', 'gobp', thresh = 0.5, genesets = genesets, data = null.y)
```

In each simulation we select one "active" gene set. Gene set genes have a probability of $p_{in} = 0.5$ of appearing the the gene list. Genes out of the gene set have a probability of $p_{out}=0.01$ of appearing in the gene list. We repeat this simulation for 3 gene sets of size ~50, 100, 300. 

Clearly, gene set overlap leads to dramatic inflation of non active gene set p-values.

```{r, aspect=0.25}
par(mfrow=c(1,4))
gaston::qqplot.pvalues(ora_null_res$ora[[1]]$pHypergeometric, main='Null Simulation')

i = 1
title = paste0(active_gene_sets[i], ', n=', ngenes[active_gene_sets[i]])
gaston::qqplot.pvalues(ora_res$ora[[i]]$pHypergeometric, main=title)

i = 2
title = paste0(active_gene_sets[i], ', n=', ngenes[active_gene_sets[i]])
gaston::qqplot.pvalues(ora_res$ora[[i]]$pHypergeometric, main=title)

i = 3
title = paste0(active_gene_sets[i], ', n=', ngenes[active_gene_sets[i]])
gaston::qqplot.pvalues(ora_res$ora[[i]]$pHypergeometric, main=title)
```


```{r}
res <- ora_res %>%
  unnest(ora) %>%
  left_join(genesets$gobp$geneSet$geneSetDes)

res %>% 
  filter(experiment == active_gene_sets[3]) %>%
  arrange(pHypergeometric) %>% 
  select(geneSet, description, overlap, geneSetSize, pHypergeometric) %>%
  head(10) %>%
  xtable::xtable(type='latex')
```


### Multiple enriched gene sets

```{r aspect=0.8}
x3 <- pmin(Matrix::rowSums(X[, active_gene_sets]), 1)
y3 <- tibble(
  ENTREZID=rownames(X),
  beta=1,
  threshold.on=sim_from_gene_set(x3, 0.5, 0.01))

y[['Y3']] <- y3
ora_y3 <- do_ora('Y3', 'gobp', 0.5, genesets = genesets, data=y)

plotdat <- ora_y3$ora[[1]] %>%
  mutate(active = geneSet %in% active_gene_sets) 


plotdat %>%
  mutate(padj = p.adjust(pHypergeometric, method = 'BH')) %>%
  mutate(reject = padj < 0.05) %>%
  ggplot(aes(x=log10(oddsRatio), y=-log10(pHypergeometric), color=reject, size=active, shape=active)) +
  geom_point() +
  scale_color_manual(values=c("FALSE"="grey","TRUE"="red")) +
  scale_size_manual(values=c('FALSE' = 1, 'TRUE'= 3)) +
  scale_shape_manual(values=c('FALSE' = 1, 'TRUE'= 17)) +
  labs(x = 'log10(Odds Ratio)', y = '-log10(p)')
```

```{r}
ora_y3$ora[[1]] %>%
  arrange(pHypergeometric) %>%
  mutate(rank = row_number()) %>%
  filter(geneSet %in% active_gene_sets) %>%
  left_join(genesets$gobp$geneSet$geneSetDes) %>%
  select(geneSet, description, rank, pHypergeometric) %>%
  xtable::xtable(type='latex')
```

## Stepwise regression

```{r}
u <- gseasusie::prep_binary_data(genesets$gobp, y3, 0.5)

y <- u$y
x3 <- u$X[, active_gene_sets[3]]
x2 <- u$X[, active_gene_sets[2]]
x1 <- u$X[, active_gene_sets[1]]

library(tictoc)
tic()
m1p <- cache_rds({map_dbl(
  colnames(u$X),
  ~tail(summary(glm(y ~ u$X[,.x], family='binomial'))$coef[, "Pr(>|z|)"], 1)
)}, file='forwardstepwise1')
names(m1p) <- colnames(u$X)
toc()

M1 <- glm(y ~ x3, family='binomial')
p1 <- predict.glm(M1) - M1$coefficients[1]

### 
tic()
m2p <- cache_rds({map_dbl(
  colnames(u$X),
  ~tail(summary(glm(y ~ u$X[,.x] + offset(p1), family='binomial'))$coef[, "Pr(>|z|)"], 1)
)}, file='forwardstepwise2')
names(m2p) <- colnames(u$X)
toc()

M2 <- glm(y ~ x2 + offset(p1), family='binomial')
p2 <- predict.glm(M2) - M2$coefficients[1]

###
tic()
m3p <- cache_rds({map_dbl(
  colnames(u$X),
  ~tail(summary(glm(y ~ u$X[,.x] + offset(p2), family='binomial'))$coef[, "Pr(>|z|)"], 1)
)}, file='forwardstepwise3')
names(m3p) <- colnames(u$X)
toc()

M3 <- glm(y ~ x1 + offset(p2), family='binomial')
p3 <- predict.glm(M3) - M3$coefficients[1]


###
tic()
m4p <- map_dbl(
  colnames(u$X),
  ~tail(summary(glm(y ~ u$X[,.x] + offset(p3), family='binomial'))$coef[, "Pr(>|z|)"], 1)
)
names(m4p) <- colnames(u$X)
toc()
```

## Correct selection path regression

```{r aspect=0.25}
par(mfrow=c(1, 4))

plot(-log10(m1p), col='white', main='Step 1')
abline(v=which(names(m3p) %in% active_gene_sets), col='grey', lty='dotted')
points(-log10(m1p))
abline(
  v=which(names(m3p) == active_gene_sets[3]),
  col='red'
)

plot(-log10(m2p), col='white', main='Step 2')
abline(v=which(names(m3p) %in% active_gene_sets), col='grey', lty='dotted')
points(-log10(m2p))
abline(
  v=which(names(m2p) == active_gene_sets[2]),
  col='red'
)

plot(-log10(m3p), col='white', main='Step 3')
abline(v=which(names(m3p) %in% active_gene_sets), col='grey', lty='dotted')
points(-log10(m3p))
abline(
  v=which(names(m3p) == active_gene_sets[1]),
  col='red'
)

plot(-log10(m4p), main='Step 4')
abline(v=which(names(m3p) %in% active_gene_sets), col='grey', lty='dotted')
```

### Actual selection path

```{r}
sort(m1p) %>% head()
sort(m2p) %>% head()
sort(m3p) %>% head()
active_gene_sets
```

### Logistic vs Linear

```{r}
susie_y3 <- gseasusie::do_logistic_susie('Y3', 'gobp', 0.5, genesets = genesets, data=list(Y3=y3))
linear_susie_y3 <- gseasusie::do_linear_susie('Y3', 'gobp', 0.5, genesets = genesets, data=list(Y3=y3))
```


```{r aspect=0.5}
par(mfrow=c(1,2))
susieR::susie_plot(susie_y3$fit[[1]], 'PIP', main='Logistic')
abline(v=which(names(m3p) %in% active_gene_sets), col='grey', lty='dotted')

susieR::susie_plot(linear_susie_y3$fit[[1]], 'PIP', main='Linear')
abline(v=which(names(m3p) %in% active_gene_sets), col='grey', lty='dotted')
```

