---
title: "scale_v_standardize_v_raw"
author: "karltayeb"
date: "2022-04-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

The way VEB.boost handles the intercept is better than estimating a fixed intercept.
Here we try:
* scaled X
* centered X
* untrasnformed X


## Why does centering help?

For a univariate regression we can think of centering/shifting $x$ as just shifting the intercept.

That is the regression coefficients we estimate using a centered version of $x$,
$\tilde x = x - \bar x$ will be 

$$
\tilde\beta_0 = \hat\beta_0 - \tilde\beta_1 \bar x\\
\tilde \beta_1 = \hat \beta_1
$$

Where $(\hat\beta_0, \hat\beta_1)$ are the coefficients estimated from the un-centered $x$.

Although a bit harder to visualize, the same holds true for multivariate $X$, where we center each column of $X$, the offsets get absorbed into the intercept in a deterministic way.

So why would centering have an impact? I suppose when the model treats the slope and intercept differently there could be a difference. That is if we have a flat prior on the intercept and a regularized coefficient? But that doesn't quite line up for me since the coefficient estimates should be the same?

```{r load.data}
library(tidyverse)
source('code/load_gene_sets.R')
source('code/load_data.R')

genesets <- load_gene_sets()
data <- load_sc_pbmc()
```

```{r fit.l1}
library(tictoc)
source('code/logistic_susie_vb.R')
source('code/enrichment_pipeline.R')

do_logistic_susie2 = partial(do_logistic_susie, genesets=genesets, data=data)
do_ora2 = partial(do_ora, genesets=genesets, data=data)

tic('Logistic SuSiE Sparse X')
susie.args = list(L=1, standardize=F, center=F, verbose=T)
l1 <- do_logistic_susie2('CD19+ B','gomf',1e-4,susie.args = susie.args)
toc()

tic('Logistic SuSiE Centered X')
susie.args = list(L=1, standardize=F, center=T, verbose=T)
l1.centered <- do_logistic_susie2('CD19+ B','gomf',1e-4,susie.args = susie.args)
toc()

tic('Logistic SuSiE Standardized X')
susie.args = list(L=1, standardize=T, center=T, verbose=T)
l1.standardized <- do_logistic_susie2('CD19+ B','gomf',1e-4,susie.args = susie.args)
toc()

l1.combined <- rbind(l1, l1.centered, l1.standardized)
l1.combined <- l1.combined %>%
  mutate(X = c('X', 'X.centered', 'X.standardized'))
ora = do_ora2('CD19+ B','gomf',1e-4)
```

```{r l10}
tic('Logistic SuSiE Sparse X')
susie.args = list(L=10, standardize=F, center=F, verbose=T)
l10 <- do_logistic_susie2('CD19+ B','gomf',1e-4,susie.args = susie.args)
toc()

tic('Logistic SuSiE Standardized X')
susie.args = list(L=10, standardize=F, center=T, verbose=T)
l10.centered <- do_logistic_susie2('CD19+ B','gomf',1e-4,susie.args = susie.args)
toc()

tic('Logistic SuSiE Standardized X')
susie.args = list(L=10, standardize=T, center=T, verbose=T)
l10.standardized <- do_logistic_susie2('CD19+ B','gomf',1e-4,susie.args = susie.args)
toc()

l10.combined <- rbind(l10, l10.centered, l10.standardized)
l10.combined <- l10.combined %>%
  mutate(X = c('X', 'X.centered', 'X.standardized'))
```


```{r}
get_cs_summary_condensed = function(fit){
  fit %>%
    get_credible_set_summary() %>%
    group_by(geneSet) %>%
    arrange(desc(alpha)) %>%
    filter(row_number() == 1)
}

fits <- l1.combined

# gene set summary
res <- fits %>%
  left_join(ora) %>%
  mutate(
    gs_summary = map(fit, get_gene_set_summary),
    cs_summary = map(fit, get_cs_summary_condensed),
    res = map2(gs_summary, cs_summary, ~ left_join(.x, .y, by='geneSet')),
    res = map2(res, ora, ~ left_join(.x, .y))
  ) %>%
  select(-c(fit, susie.args, ora, gs_summary, cs_summary)) %>%
  unnest(res)

# gene set summary
res.l10 <- l10.combined %>%
  left_join(ora) %>%
  mutate(
    gs_summary = map(fit, get_gene_set_summary),
    cs_summary = map(fit, get_cs_summary_condensed),
    res = map2(gs_summary, cs_summary, ~ left_join(.x, .y, by='geneSet')),
    res = map2(res, ora, ~ left_join(.x, .y))
  ) %>%
  select(-c(fit, susie.args, ora, gs_summary, cs_summary)) %>%
  unnest(res)
```

```{r}
plot(l1$fit[[1]]$mu, l1.centered$fit[[1]]$mu)
```


```{r susie.plots}
library(susieR)
par(mfrow=c(1,3))
susie_plot(l1$fit[[1]], 'PIP')
susie_plot(l1.centered$fit[[1]], 'PIP')
susie_plot(l1.standardized$fit[[1]], 'PIP')
```

```{r}
l1.combined$fit[[1]]$set$cs
l1.combined$fit[[2]]$set$cs
l1.combined$fit[[3]]$set$cs
```




```{r volcano.plots}
get_ora_enrichments = function(tbl){
   tbl %>% mutate(
    padj = p.adjust(pFishersExact),
    result = case_when(
      padj < 0.05 & oddsRatio < 1 ~ 'depleted',
      padj < 0.05 & oddsRatio > 1 ~ 'enriched',
      TRUE ~ 'not significant'
    )
  )
}

do.volcano = function(res){
  res %>% 
    get_ora_enrichments %>%
    ggplot(aes(x=log10(oddsRatio), y=-log10(pFishersExact), color=result)) +
    geom_point() +
    geom_point(
      res %>% filter(in_cs, active_cs), 
      mapping=aes(x=log10(oddsRatio), y=-log10(pFishersExact)),
      color='black', pch=21, size=5) +
    scale_color_manual(values = c('depleted' = 'coral',
                                  'enriched' = 'dodgerblue',
                                  'not significant' = 'grey'))
}

do.volcano(res) + facet_wrap('X')
do.volcano(res.l10) + facet_wrap('X')
```





