---
title: "Single Cell PBMC Example: Shared intercept explains difference with univariate enrichment tests"
author: "karltayeb"
date: "2022-04-03"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

## Summary

When we fit logistic SuSiE, with $L=1$ (SER) or $L=10$ we often see that the gene sets selected in the credible sets are not the gene sets with the highest marginal significance (by hypergeometric test p-value and univariate logistic regression p-value from `glm`)

We explore if this is driven by (1) shrinkage of the effects or (2) the shared intercept term in the SER. If you go back and fit the univariate logistic regression with the intercept fixed to the SER estimate, we find that SER and univariate logistic regression agree.

We can also see that there are differences between the hypergeometric p-values and the logistic regression p-values. But these are explained by negative effect sizes (depletion) in the logistic regression and small, highly observed gene sets (which are an edge case for the hypergeometric test, I suppose).

Finally, we can't rule out the variational approximation causing some differences, but we don't need to blame the approximation in this case.

```{r message=FALSE, results='hide'}
library(GSEABenchmarkeR)
library(EnrichmentBrowser)
library(tidyverse)
library(susieR)
library(DT)
library(kableExtra)

source('code/load_gene_sets.R')
source('code/utils.R')
source('code/logistic_susie_vb.R')
source('code/logistic_susie_veb_boost.R')
source('code/latent_logistic_susie.R')
```

```{r geneset.setup, message=FALSE, results='hide'}
gs_list <- WebGestaltR::listGeneSet()
gobp <- loadGeneSetX('geneontology_Biological_Process', min.size=50)  # just huge number of gene sets
gobp_nr <- loadGeneSetX('geneontology_Biological_Process_noRedundant', min.size=1)
gomf <- loadGeneSetX('geneontology_Molecular_Function', min.size=1)
kegg <- loadGeneSetX('pathway_KEGG', min.size=1)
reactome <- loadGeneSetX('pathway_Reactome', min.size=1)
wikipathway_cancer <- loadGeneSetX('pathway_Wikipathway_cancer', min.size=1)
wikipathway <- loadGeneSetX('pathway_Wikipathway', min.size=1)

genesets <- list(
  gobp=gobp,
  gobp_nr=gobp_nr,
  gomf=gomf,
  kegg=kegg,
  reactome=reactome,
  wikipathway_cancer=wikipathway_cancer,
  wikipathway=wikipathway
)
```

```{r data.setup}
load('data/pbmc-purified/deseq2-pbmc-purified.RData')
```


```{r driver}
convert_labels <- function(y, from='SYMBOL', to='ENTREZID'){
  hs <- org.Hs.eg.db::org.Hs.eg.db
  gene_symbols <- names(y)
  symbol2entrez <- AnnotationDbi::select(
    hs, keys=gene_symbols, columns=c(to, from), keytype = from)
  symbol2entrez <- symbol2entrez[!duplicated(symbol2entrez[[from]]),]
  symbol2entrez <- symbol2entrez[!is.na(symbol2entrez[[to]]),]
  symbol2entrez <- symbol2entrez[!is.na(symbol2entrez[[from]]),]
  rownames(symbol2entrez) <- symbol2entrez[[from]]
  ysub <- y[names(y) %in% symbol2entrez[[from]]]
  names(ysub) <- symbol2entrez[names(ysub),][[to]]
  return(ysub)
}
convert_labels = partial(convert_labels, from='ENSEMBL')

#' take gene level results and put them in a standard format
#' a named list with ENTREZID names and gene level stats
#' target_col is the column to extract
#' from is the source gene label to convert to ENTREZID
clean_gene_list = function(data, target_col){
  target_col = sym(target_col)
  data %>%
    as.data.frame %>%
    rownames_to_column('gene') %>%
    dplyr::select(gene, !!target_col) %>%
    filter(!is.na(!!target_col)) %>%
    mutate(y = !!target_col) %>%
    select(gene, y) %>%
    tibble2namedlist %>%
    convert_labels()
}
clean_gene_list = partial(clean_gene_list, target_col='padj')

#' makes a binary list from table like data
get_y = function(data, thresh=1e-4){
  data %>%
    clean_gene_list() %>%
    {
      y <- as.integer(. < thresh)
      names(y) <- names(.)
      y
    }
}
get_y(deseq$`CD19+ B`, 1e-40) %>% mean()

#' fit logistic regression to each gene set individually
do_marginal_logistic_regression = function(db,
                                           celltype,
                                           thresh,
                                           glm.args = list(family='binomial')){
  gs <- genesets[[db]]
  data <- deseq[[celltype]]
  y <- get_y(data, thresh)
  u <- process_input(gs$X, y)  # subset to common genes
  
  n <- dim(u$X)[1]  # number of genes
  p <- dim(u$X)[2]  # number of gene sets
  f <- exec(partial, glm, !!!glm.args)
  library(tictoc)
  tic()
  marginal.fit <- purrr::map(1:p, ~ f(u$y ~ u$X[,.x]))
  toc()
  names(marginal.fit) <- colnames(u$X)[1:p]

  return(list(
    fit = marginal.fit,
    db = db, celltype = celltype, thresh = thresh))
}

#' fit logistic susie, and hypergeometric test
do_logistic_susie = function(db, celltype, thresh, susie.args=NULL){
  gs <- genesets[[db]]
  data <- deseq[[celltype]]
  y <- get_y(data, thresh)
  u <- process_input(gs$X, y)  # subset to common genes
  
  if(is.null(susie.args)){
    susie.args = list(
      L=10, init.intercept=0, verbose=1, maxit=100, standardize=TRUE)
  }
  vb.fit <- exec(logistic.susie, u$X, u$y, !!!susie.args)

  #' hypergeometric test
  ora <- tibble(
    geneSet = colnames(u$X),
    geneListSize = sum(u$y),
    geneSetSize = colSums(u$X),
    overlap = (u$y %*% u$X)[1,],
    nGenes = length(u$y),
    propInList = overlap / geneListSize,
    propInSet = overlap / geneSetSize,
    oddsRatio = (overlap / (geneListSize - overlap)) / (
      (geneSetSize - overlap) / (nGenes - geneSetSize + overlap)),
    pValueHypergeometric = phyper(
      overlap-1, geneListSize, nGenes - geneListSize, geneSetSize, lower.tail= FALSE),
    nl10p = -log10(pValueHypergeometric),
    db = db,
    celltype = celltype,
    thresh = thresh
  ) %>% 
  left_join(gs$geneSet$geneSetDes)

  return(list(
    fit = vb.fit,
    ora = ora,
    db = db, celltype = celltype, thresh = thresh))
}

#' fit logistic susie, and hypergeometric test
do_logistic_susie_veb_boost = function(db, celltype, thresh, susie.args=NULL){
  gs <- genesets[[db]]
  data <- deseq[[celltype]]
  y <- get_y(data, thresh)
  u <- process_input(gs$X, y)  # subset to common genes
  
  X <- u$X
  dim(X[,(colSums(X) > 0 & colSums(X) < dim(X)[1])])
  dim(X)
  if(is.null(susie.args)){
    susie.args = list(L=10)
  }
  vb.fit <- exec(logistic.susie.veb.boost, u$X, u$y, !!!susie.args)

  veb.fit <- VEB.Boost::veb_boost_stumps(
    u$X, u$y, k=2,
    family='binomial',
    include_stumps = F,
    growTree=F,
    changeToConstant=F
  )
  
  sum(colSums(u$X) == 12210)
  
  veb.fit$X
  
  alpha <- t(do.call(cbind, lapply(veb.fit$leaves, function(x) x$currentFit$alpha)))
  mu <- t(do.call(cbind, lapply(veb.fit$leaves, function(x) x$currentFit$mu)))
  mu2 <- t(do.call(cbind, lapply(veb.fit$leaves, function(x) x$currentFit$mu2)))
  elbo <- veb.fit$ELBO_progress[[2]]
  res <- list(alpha=alpha, mu=mu, mu2=mu2, elbo=elbo, veb.fit=veb.fit)
  
  class(res) <- 'susie'
  colnames(res$alpha) <- colnames(u$X)
  colnames(res$mu) <- colnames(X)
  res$pip <- susieR::susie_get_pip(res)
  names(res$pip) <- colnames(X)
  res$sets <- susieR::susie_get_cs(res, X=X)
 
  #' hypergeometric test
  ora <- tibble(
    geneSet = colnames(u$X),
    geneListSize = sum(u$y),
    geneSetSize = colSums(u$X),
    overlap = (u$y %*% u$X)[1,],
    nGenes = length(u$y),
    propInList = overlap / geneListSize,
    propInSet = overlap / geneSetSize,
    oddsRatio = (overlap / (geneListSize - overlap)) / (
      (geneSetSize - overlap) / (nGenes - geneSetSize + overlap)),
    pValueHypergeometric = phyper(
      overlap-1, geneListSize, nGenes - geneListSize, geneSetSize, lower.tail= FALSE),
    nl10p = -log10(pValueHypergeometric),
    db = db,
    celltype = celltype,
    thresh = thresh
  ) %>% 
  left_join(gs$geneSet$geneSetDes)

  return(list(
    fit = vb.fit,
    ora = ora,
    db = db, celltype = celltype, thresh = thresh))
}

get_credible_set_summary = function(res){
  gs <- genesets[[res$db]]
  data <- deseq[[res$celltype]]

  #' report top 50 elements in cs
  credible.set.summary <- t(res$fit$alpha) %>%
    data.frame() %>%
    rownames_to_column(var='geneSet') %>%
    rename_with(~str_replace(., 'X', 'L')) %>%
    rename(L1 = 2) %>%  # rename deals with L=1 case
    pivot_longer(starts_with('L'), names_to='component', values_to = 'alpha') %>%
    arrange(component, desc(alpha)) %>%
    dplyr::group_by(component) %>%
    filter(row_number() < 50) %>%
    mutate(alpha_rank = row_number(), cumalpha = c(0, head(cumsum(alpha), -1))) %>%
    mutate(in_cs = cumalpha < 0.95) %>%
    mutate(active_cs = component %in% names(res$fit$sets$cs)) %>%
    left_join(res$ora) %>%
    left_join(gs$geneSet$geneSetDes)

  return(credible.set.summary)
}

get_gene_set_summary = function(res){
  gs <- genesets[[res$db]]
  #' map each gene set to the component with top alpha
  #' report pip
  res$fit$pip %>% 
    as_tibble(rownames='geneSet') %>%
    rename(pip=value) %>%
    mutate(beta=colSums(res$fit$alpha * res$fit$mu)) %>%
    left_join(res$ora) %>%
    left_join(gs$geneSet$geneSetDes)
}
```

```{r plot.function}
do.or.plot = function(res, 
                      ct=NULL, 
                      g=NULL, 
                      x='oddsRatio', 
                      s='geneSetSize', 
                      c='in_cs'){
  col_sym = sym(x)
  gene.set.order <- res$geneset.summary %>% 
    filter(if(!is.null(ct)) celltype == ct else TRUE) %>%
    filter(if(!is.null(g)) db == g else TRUE) %>%
    arrange(db, !!col_sym) %>%
    .$geneSet %>% 
    unique()
  
  size_sym = sym(s)
  color_sym = sym(c)
  plot <- res$cs.summary %>%
    filter(if(!is.null(ct)) celltype == ct else TRUE) %>%
    filter(if(!is.null(g)) db == g else TRUE) %>%
    filter(active_cs, alpha_rank <= 10) %>%
    mutate(geneSet=factor(geneSet, levels=gene.set.order)) %>%
    ggplot(aes(x=!!col_sym, y=geneSet, color=!!color_sym, size=!!size_sym)) +
    geom_point() +
    facet_wrap(vars(component), scale='free')
  return(plot)
}
```


```{r fit.logistic.regression}
#' fit logistic regression to each gene set individually
summarize_fit = function(glm.fitted){
  coef = summary(glm.fitted)$coefficients
  return(list(
    logistic.intercept = coef[1, 1],
    logistic.intercept.se = coef[1,2],
    logistic.effect = coef[2,1],
    logistic.effect.se = coef[2,2],
    logistic.p.value = coef[2,4],
    logistic.regression.nl10p = -log10(coef[2,4] + 1e-20),
    logistic.aic = glm.fitted$aic
  ))
}

summarize_fit = function(glm.fitted){
  coef = summary(glm.fitted)$coefficients
  return(list(
    coef = coef,
    logistic.aic = glm.fitted$aic
  ))
}

do_marginal_logistic_regression = function(db,
                                           celltype,
                                           thresh,
                                           glm.args = list(family='binomial'), intercept=T){
  gs <- genesets[[db]]
  data <- deseq[[celltype]]
  y <- get_y(data, thresh)
  u <- process_input(gs$X, y)  # subset to common genes
  
  n <- dim(u$X)[1]  # number of genes
  p <- dim(u$X)[2]  # number of gene sets
  
  f <- exec(partial, glm, !!!glm.args)
  library(tictoc)
  tic()
  if(intercept){
    marginal.fit <- purrr::map(1:p, ~ possibly(summarize_fit, NULL)(f(u$y ~ u$X[,.x])))
  } else {
    marginal.fit <- purrr::map(1:p, ~ possibly(summarize_fit, NULL)(f(u$y ~ u$X[,.x] - 1)))
  }
  names(marginal.fit) <- colnames(u$X)[1:p]
  toc()

  marginal <- tibble(
    geneSet = names(marginal.fit),
    fit = marginal.fit) %>%
    unnest_wider(fit) %>%
    mutate(
      db = db,
      celltype = celltype,
      thresh = thresh
    )
  return(marginal)
}

marginal_tbl <- xfun::cache_rds({
  do_marginal_logistic_regression('gomf', 'CD19+ B', 1e-4)},
  dir = 'cache/single_cell_pbmc_l1/',
  file='fit.logistic.regression')
```


```{r fit.models}
# L =10
susie.args = list(L=10, standardize=F, verbose=T)
susie.l10 = do_logistic_susie('gomf', 'CD19+ B', 1e-4, susie.args)
res.l10 = list(
  cs.summary = get_credible_set_summary(susie.l10),
  geneset.summary = get_gene_set_summary(susie.l10)
)

# L =1
susie.args <- list(L=1, standardize=F, verbose=T)
susie.l1 <- do_logistic_susie('gomf', 'CD19+ B', 1e-4, susie.args)
res.l1 <- list(
  cs.summary = get_credible_set_summary(susie.l1),
  geneset.summary = get_gene_set_summary(susie.l1)
)

# L =1 "flat" prior
susie.args <- list(L=1, estimate_prior_variance = F, V= 1000, standardize=F, verbose=T)
susie.l1.flat <- do_logistic_susie('gomf', 'CD19+ B', 1e-4, susie.args)
res.l1.flat <- list(
  cs.summary = get_credible_set_summary(susie.l1.flat),
  geneset.summary = get_gene_set_summary(susie.l1.flat)
)

# L = 1 "flat" prior
susie.args <- list(L=1)
susie.l1.veb <- do_logistic_susie_veb_boost('gomf', 'CD19+ B', 1e-4, susie.args)
res.l1.veb <- list(
  cs.summary = get_credible_set_summary(susie.l1.veb),
  geneset.summary = get_gene_set_summary(susie.l1.veb)
)

# L = 10 "flat" prior
susie.args <- list(L=1)
susie.l10.veb <- do_logistic_susie_veb_boost('gomf', 'CD19+ B', 1e-4)
res.l10.veb <- list(
  cs.summary = get_credible_set_summary(susie.l10.veb),
  geneset.summary = get_gene_set_summary(susie.l10.veb)
)
```


```{r fit.fixed.intercept}
# fixed intercept model
intercept <- susie.l1$fit$intercept
n <- susie.l1$fit$dat$y %>% length()

marginal.fixed.intercept <- xfun::cache_rds({
  glm.args <- list(family=binomial(), offset=rep(intercept, n))
  do_marginal_logistic_regression(
    'gomf', 'CD19+ B', 1e-4, glm.args = glm.args, intercept=F)},
  dir = 'cache/single_cell_pbmc_l1/',
  file='fit.logistic.regression.fixed.intercept')


marginal.fixed.intercept <-
  marginal.fixed.intercept %>%
  mutate(
    nl10p.fixed.intercept = -log10(marginal.fixed.intercept$coef[, 4]),
    loglik.fixed.intercept = -0.5 *(logistic.aic - 4)
  )

marginal <- 
  marginal_tbl %>%
  filter(row_number() > 1) %>%
  rowwise() %>%
  mutate(
    nl10p.beta = -log10(tail(coef[,4], 1))
  ) %>% ungroup() %>%
  mutate(
    loglik.beta = -0.5 *(logistic.aic - 4)
  ) %>%
  select(-c(coef, logistic.aic)) %>%
  left_join(marginal.fixed.intercept %>% select(-coef))



p1 <- marginal %>%
  ggplot(aes(x=loglik.beta, y=loglik.fixed.intercept)) +
  geom_point()
p2 <- marginal %>%
  ggplot(aes(x=nl10p.beta, y=nl10p.fixed.intercept)) +
  geom_point()
cowplot::plot_grid(p1, p2)
```


```{r update.results}
res.l10$cs.summary <-
  res.l10$cs.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )

res.l10$geneset.summary <-
  res.l10$geneset.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )

res.l1$cs.summary <-
  res.l1$cs.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )

res.l1$geneset.summary <-
  res.l1$geneset.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )

res.l1.flat$cs.summary <-
  res.l1.flat$cs.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )
res.l1.flat$geneset.summary <-
  res.l1.flat$geneset.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )

res.l1.veb$cs.summary <-
  res.l1.veb$cs.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )
res.l1.veb$geneset.summary <-
  res.l1.veb$geneset.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )

res.l10.veb$cs.summary <-
  res.l10.veb$cs.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )
res.l10.veb$geneset.summary <-
  res.l10.veb$geneset.summary %>%
  left_join(marginal, by = c('geneSet', 'db', 'celltype')) %>%
  mutate(
    minor.gene.frequency = geneSetSize / nGenes,
    minor.gene.frequency = pmin(1 - minor.gene.frequency, minor.gene.frequency)
  )
```

### $L=10$

The gene set selected by logistic susie with $L=10$ are not the gene sets with the smallest p-value from a hypergeometric test. This is partially remedied by ranking by the p-values ($L2$ get's fixed) from logistic regression. If we fit the logistic regression with a fixed intercept we also recover $L1$. We don't get the other components, but it's possible the effects estimates are not independent. Let's look at the SER case.

```{r susie.plots}
do.or.plot(res.l10, x='nl10p', s='minor.gene.frequency')
do.or.plot(res.l10, x='nl10p.beta', s='minor.gene.frequency')
do.or.plot(res.l10, x='nl10p.fixed.intercept', s='minor.gene.frequency')
```


### $L=1$

Consistent with $L=10$ only when we fix the regression intercept do we get pvalues that agree with the PIPs.

```{r ser.plots}
p1 <- do.or.plot(res.l1, x='nl10p', s='minor.gene.frequency')
p2 <-do.or.plot(res.l1, x='nl10p.beta', s='minor.gene.frequency')
p3 <-do.or.plot(res.l1, x='nl10p.fixed.intercept', s='minor.gene.frequency')

cowplot::plot_grid(p1, p2, p3)
```



### $L=1$ Flat

I also wanted to see if shrinkage on the effect estimates was causing any differences.
The enrichment plot looks remarkable similar but if we plot `mu` from SER we see that the effects are shrunk (although that's a pretty funky shape-- it looks like smaller gene sets experience stronger shrinkage which makes sense).

```{r ser.flat.plots}
p1 <- do.or.plot(res.l1.flat, x='nl10p', s='minor.gene.frequency')
p2 <-do.or.plot(res.l1.flat, x='nl10p.beta', s='minor.gene.frequency')
p3 <-do.or.plot(res.l1.flat, x='nl10p.fixed.intercept', s='minor.gene.frequency')

cowplot::plot_grid(p1, p2, p3, nrow = 1)
```

```{r check.shrinkage}
sizes <- colSums(genesets$gomf$X[, colnames(susie.l1$fit$mu)])
tbl <- tibble(
  flat.mu=susie.l1.flat$fit$mu[1,],
  mu=susie.l1$fit$mu[1,],
  size = sizes)

tbl %>% 
  ggplot(aes(y=mu, x=flat.mu, color=log(size))) +
  geom_point()
```


### $L=1$ VEB.Boost (estimate intercept per effect)

```{r veb.1.plots}
p1 <- do.or.plot(res.l1.veb, x='nl10p', s='minor.gene.frequency')
p2 <-do.or.plot(res.l1.veb, x='nl10p.beta', s='minor.gene.frequency')
p3 <-do.or.plot(res.l1.veb, x='nl10p.fixed.intercept', s='minor.gene.frequency')

cowplot::plot_grid(p1, p2, p3, nrow = 1)
```


### $L=10$ VEB.Boost (estimate intercept per effect)

```{r sveb.10.plots}
p1 <- do.or.plot(res.l10.veb, x='nl10p', s='minor.gene.frequency')
p2 <-do.or.plot(res.l10.veb, x='nl10p.beta', s='minor.gene.frequency')
p3 <-do.or.plot(res.l10.veb, x='nl10p.fixed.intercept', s='minor.gene.frequency')

cowplot::plot_grid(p1, p2, p3, ncol = 1)
```

### Some other plots

log PIPs vs nl10p from hypergeometric test, logitic regression, and fixed intercept regression

```{r pip.v.nl10p}
p1 <- res.l1$geneset.summary %>% ggplot(aes(x=log(pip), y=nl10p)) + geom_point()
p2 <- res.l1$geneset.summary %>% ggplot(aes(x=log(pip), y=nl10p.beta)) + geom_point()
p3 <- res.l1$geneset.summary %>% ggplot(aes(x=log(pip), y=nl10p.fixed.intercept)) + geom_point()
cowplot::plot_grid(p1, p2, p3, nrow = 1)
```


There are some discrepancies between the pvalues from different methods.
One class of differences are where SuSiE estimates a negative effect. This depletion is not detected by the one-sided hypergeometric test.
For some small, highly observed, gene sets, the hypergeometric test is giving a much stronger pvalue than the logistic regression.

```{r hyper.v.logistic}
p1 <- res.l1$geneset.summary %>% ggplot(aes(x=nl10p.beta, y=nl10p, color=(beta > 0))) + geom_point()
p2 <- res.l1$geneset.summary %>%
  mutate(geneSetSize = case_when(
      geneSetSize < 10 ~ '<10',
      geneSetSize < 50 ~ '10-50',
      geneSetSize < 100 ~ '50-100',
      TRUE ~ '>100')) %>%
  ggplot(aes(x=nl10p.beta, y=nl10p, color=factor(geneSetSize, levels = c('<10', '10-50', '50-100', '>100')))) +
  geom_point() +
  scale_color_discrete('Gene set size')
cowplot::plot_grid(p1, p2, nrow = 1)

```

```{r}
knitr::knit_exit()
```
